### 2023-07-08 manuscript_english.tex
Adding data

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels},
  pdfauthor={Norma Hernandez, Angel Robles Nathan Upham},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Suitability prevalence area index in late quaternary explains
genetic diversity in Tassel eared Squirrels}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018).

Our results reveal that fitness prevalence corresponds to the fixation
index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

The current distributions of species do not always correspond to their
historical distributions over evolutionarily significant periods. This
is because environmental conditions are not static over time; species
tend to distribute where conditions are most favorable.

It follows from this relationship between environment and species
distribution that population size also varies with time. This variation
in population size is related to the effective population size \(N_e\)
(Karlin 1968), so that an index reflecting changes in environmental
conditions in geography might be expected to be related to \(N_e\) and
thus to indicators of population structure such as the fixation index
\(F_{st}\). Thus, it is possible to relate patterns of changes in the
distribution of species to the genetic structure of their populations
using a statistical model that explains this relationship.

With this approach we can predict the geographic pattern of population
structure from environmental information. This approach is strongly
driven by advances in currently available climate simulations (Leonardi
et al. 2023; Krapp et al. 2021), as well as next generation sequencing
data, and supported by both ecological niche (Thorup et al. 2021;
Nogués-Bravo 2009) and population genetics theories (Lira-Noriega and
Manthey 2014).

In this work we propose a method to find the Suitability Prevalence Area
(SPA) as an index with a double purpose: 1) to find endemic areas to
delimit the historical distribution of the species and 2) to explain the
patterns of genetic diversity.

To obtain the SPA, we performed a historical reconstruction of the
geographical range back to 120 000 BC at 2 000 year intervals and
recorded the environmental suitability at each site.

Subsequently, we delimited historical endemic areas to locations where
the prevalence of suitability remained at 90\% during this period. From
the fixation index calculated with respect to populations in the
historical endemic areas, a statistical model of the fixation index as a
function of SPA was performed. With this statistical model, the fixation
index values were projected to the current distribution to obtain a map
with the geographic pattern of this index.

\hypertarget{theoretical-basis}{%
\subsubsection{Theoretical basis}\label{theoretical-basis}}

\hypertarget{environment-fitness-and-suitability}{%
\paragraph{Environment, fitness and
suitability}\label{environment-fitness-and-suitability}}

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. However, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
\caption{Hypothetical niche in environmental space. An arbitrary point
moves within the ambient space entering and exiting the volume (niche)
where the intrinsic growth rate is positive.}
\end{figure}

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and that for each trajectory the fitness function
changes and therefore the suitability function changes but not
accordingly. in a homogeneous way, if not more so that each site
presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)).

Es decir:

\[
N_i \propto  S_i(\vec{e}_i)
\] Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)

\hypertarget{genetic-variation-and-effective-population-size}{%
\paragraph{Genetic variation and effective population
size}\label{genetic-variation-and-effective-population-size}}

On the other hand, loss of heterozygosity may be related to the amount
of genetic variation present in the absence of mutation and selection,
so theoretically, one would expect a strong correlation between
effective population size and heterozygosity on the basis of population
genetics. We start from the assumption that in the absence of gene flow
between subpopulations, the rate of fixation by isolation depends on the
effective population size \(N_e\) when subpopulations diverge from a
common ancestor in \(t\) generations. Wright (1943):

\[
F_{st}(t) = 1 - e^{-kt/N_e } 
\]

where we assume, according to Nei (Nei 1986), that genetic drift within
each subpopulation causes the average heterozygosity between populations
(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)), i.e., for \(t=0\)
\(F_{st} = 0\) and for very large \(t\) the index is \(1\). that is, for
\(t=0\) \(F_{st} = 0\) and for very large \(t\) the index is \(1\). We
have included a constant \(k\) without loss of generalization.

Using
\(e(x) = \sum_{k=0}^{\infty} \frac{x^k}{k!} = 1 + x + \frac{x^2}{2} + ...\)
and leaving to first order for \(N_e\) large enough

\[
F_{st} \approx \frac{kt}{N_e} + C
\]

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
We substitute the harmonic mean into the above equation, considering
that although the number of generations is not exactly equal to the time
of the fixation index, they are expected to be very close.

\[
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
\]

\hypertarget{relationship-between-the-fixation-index-and-environmental-suitability}{%
\paragraph{Relationship between the fixation index and environmental
suitability}\label{relationship-between-the-fixation-index-and-environmental-suitability}}

We can infer locally for each population \(N_i\) in the geography expect
to be proportional to the environmental suitability for each time
considered, For simplification in the index we select a \(N_i\)
population in the geography and infer that locally, this population is
proportional to the environmental suitability in each time considered,
i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
generation time), as we describe above, so we can asume an harmonic mean
of suitabilities
\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
substitue in the previous equation:

\[
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{S_j(\vec{e_j})}
\]

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the fitness prevalence for an \(i\) site in
the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean of fitness \emph{Suitability prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

We summarize our 5 steps predictive approach in the figure 2.

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{all_figures/figure_2.png}
\caption{Realized niche in environmental space. A point dynamically
moves along different time slices}
\end{figure}

We propose a method to find the Suitability Prevalence Area (SPA) as an
index with a dual purpose: 1) to find endemic areas to delimit the
historical distribution of species and 2) to explain patterns of genetic
diversity.

The method for estimating SPA consists of a series of 4 steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past analogous climates and retain projections only for
  sites with climates analogous to the present.
\item
  Generate a harmonic mean suitability map for all time slices.
\end{enumerate}

Subsequently, one application of the SPA is to explain genetic diversity
(the fixation index) as a function of the SPA and project it to the
geography. For this it is necessary to perform a series of steps, which
consist of:

\begin{itemize}
\tightlist
\item
  Obtain georeferenced points of genetically structured populations with
  fixation index values.
\item
  Extract values from the SPA map with these points.
\item
  A linear model of the fixation index value with respect to the SPA is
  performed.
\item
  With the linear model all SPA values of suitability are interpolated.
\item
  A map is generated with all the values interpolated by the model.
\end{itemize}

To test our model we use the Aberti squirrel (\emph{Sciurus aberti}) as
a case study because:

\begin{itemize}
\tightlist
\item
  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
\item
  The subspecies are geographically isolated, since they are found in
  pine forests surrounded by desert. Therefore we can consider the
  assumption of isolation by distance (Wright 1943).
\item
  There are sufficient occurrence records for the entire range of the
  known distribution.
\item
  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
  therefore, we assume that its fundamental niche is stable over time.
\item
  There is fixation index information (Bono et al. 2018) for different
  populations of the subspecies, which is useful for using SPA as a way
  to explain genetic differentiation between populations.
\end{itemize}

\hypertarget{unique-occurrence-records}{%
\subsubsection{Unique occurrence
records}\label{unique-occurrence-records}}

We have downloaded geographic information on Sciurus aberti from the
open access platform Global Biodiversity Information Facility (GBIF).
Subsequently, we eliminated duplicate records, those without precise
coordinates and those with uncorroborated collection information. Once
the unique occurrence records were obtained, we filtered the geographic
space in order to avoid overpredictions in the spatial distribution
models caused by the agglomeration of occurrence points. The filtering
consisted of superimposing the records in the geographic space, then we
made 5 latitudinal slices of 5 degrees of separation between each one,
from latitude 20 degrees south to latitude 45 degrees north, taking as
criteria the known distribution of the species and the ecoregions of
North America, level 1 (downloaded from the US Environmental Protection
Agency, 2010). Once the geographic space was segmented, we left the same
number of records for each band (30 records per band) obtaining a total
of 150 unique records of presence.

\hypertarget{environmental-characterization}{%
\subsubsection{Environmental
characterization}\label{environmental-characterization}}

The environmental information to generate the models, from the present
and the reconstruction to the past, was obtained from Pastclim 1.2, an R
statistical software package designed to download and manipulate
paleoclimatic datasets. In our particular case we chose the Beyer at al.
Beyer, Krapp, and Manica (2020) set as it has environmental information
available up to 120 000 BE, significantly evolutionary time, at 2000
year time intervals based on Global Circulation Models HadCM3
(Singarayer and Valdes 2010) and HadAM3H (Valdes et al. 2017). The
environmental dataset has 17 bias-corrected bioclimatic variables,
reduced spatial scale and spatial resolution of 0.5°square cells.

\hypertarget{accessibility-region-textbfm.}{%
\subsubsection{\texorpdfstring{Accessibility region
(\(\textbf{M}\)).}{Accessibility region (\textbackslash textbf\{M\}).}}\label{accessibility-region-textbfm.}}

Species distribution models according to the BAM diagram (Soberon and
Peterson 2005) states that the geographic range occupied by a species
(\(\textbf{G}_o\)) is the region of appropriate assemblages in terms of
abiotic conditions. The region \(\textbf{M}\) represents in geography
areas in which the species has access due to its colonization
capabilities and the structure of geographic barriers within a specific
period of time (Soberón and Nakamura 2009).

In this study we selected a set of North American ecoregions for
modeling to delimit our Accessibility region. We took into account
ecoregions that overlap with the known distribution of the species and
also took into account neighboring ecoregions to increase the potential
accessibility in projection scenarios to the past. Thus, regions 5, 6,
9, 10, 12, 13 of the level I ecoregions (\url{https://www.epa.gov/})
were selected.

\hypertarget{niche-models}{%
\subsubsection{Niche models}\label{niche-models}}

\hypertarget{current-distribution}{%
\paragraph{Current distribution}\label{current-distribution}}

Estimating niche boundaries for species occurrence is called ecological
niche modeling (ENM), and when the emphasis is on geographic
distribution, it is known as species distribution modeling (SDM) (Guisan
and Thuiller 2005; Peterson and Soberón 2012; Saupe et al. 2012). We
model the current distribution of S. aberti using the MAXENT 3.4.4
algorithm (Phillips, Anderson, and Schapire 2006), which is a
correlative model based on the maximum entropy principle used to
estimate species distributions. The output of Maxent is a relative
occurrence rate (ROR) interpreted as a probability of habitat
suitability given observed environmental conditions by presence-only
points (Merow, Smith, and Silander Jr 2013).

The parameterization was performed with the default values of the
program with the exception of the ``extrapolate'' and ``clamping''
options to avoid artificial extrapolations in the extreme values of the
climatic variables used in the models. To calibrate the models we used
70\% of the records and the remaining 30\% to validate them.

In order to give statistical certainty and to consider uncertainty, as
well as to reduce overfitting and to have a suitable model to
extrapolate to past scenarios, 10 replications were performed for each
model using a cross-validation with the training data. In the end, the
average model of these replications was considered as the result.

We evaluated the statistical significance of the resulting models for
the present using the partial ROC test (Peterson, Papeş, and Soberón
2008), which is a modification of the ROC (receiver Operation
Characteristic) test. The results of this test are proportions (ratios)
of the area under the curve of the model with respect to a null model,
product of repetitions that allow to statistically evaluate the areas
under the curve (AUC) in relation to that expected by chance (Peterson,
Papeş, and Soberón 2008), where a value derived by chance would be \(0\)
and an acceptable value, according to the proportion of minimum omission
errors tolerated in the model, would be greater than \(1\). A different
set of test data was used on the average model of the replicates and in
this way the reduction of overfitting was guaranteed.

\hypertarget{reconstruction-to-the-past}{%
\paragraph{Reconstruction to the
past}\label{reconstruction-to-the-past}}

The average model obtained from Maxent was extrapolated to environmental
conditions into the past. Environmental layers were taken from pastclim
(Leonardi et al. 2023) in 2000 year time slices to cover a total of 120
000 years. For each scenario into the past, 10 replicates were
cross-validated and averaged.

\hypertarget{analogy-of-environmental-conditions}{%
\paragraph{Analogy of environmental
conditions}\label{analogy-of-environmental-conditions}}

To be certain that areas in the past are climatically analogous to the
present we implemented the mobility-orient parity (MOP) test. This
method generates a value from \(0\) to \(1\) according to the
environmental analogy of past conditions to the present to eliminate
geographic areas in the projections to the past outside of climates
analogous to the present (Owens et al. 2013).

\hypertarget{suitability-prevalence-area-spa}{%
\subsubsection{Suitability Prevalence Area
(SPA)}\label{suitability-prevalence-area-spa}}

Finally, to generate the SPA, both the average projection of the present
and all the average projections of the past are considered. The output
files are read in raster format.

A harmonic mean is performed as follows: For each layer the inverse of
each cell value is obtained. Then the inverse values are summed by cells
and divided by the number of scenarios. Finally, the inverse of each
cell is obtained. It is important to consider that the Maxent outputs
may contain very small values close to zero in many cells, for this
reason we added a small amount of \(0.001\) to all the values in all the
maps so that when obtaining the harmonic mean we could have an inverse
value for values close to zero without altering the biological sense of
environmental suitability.

\hypertarget{endemic-historic-area}{%
\subsubsection{Endemic historic Area}\label{endemic-historic-area}}

To calculate the historic endemic area, the SPA result is taken and a
threshold is applied to generate a binary map according to the
\(mathbf{S}\) equation. This threshold is the proportion of time that
pixel has had ideal conditions. For our study, we consider the value of
90\% since for a period of 120 000 years it is a coverage of 100 000
years, which is a significantly evolutionary time to fix genetic
patterns at the population level.

\hypertarget{fixation-index-as-a-functino-of-spa}{%
\subsubsection{Fixation index as a functino of
SPA}\label{fixation-index-as-a-functino-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

\hypertarget{fixation-index-projected-in-geography}{%
\subsubsection{Fixation index projected in
geography}\label{fixation-index-projected-in-geography}}

The linear model that presents the highest value of \(R^2\) is
considered suitable to make a geographic projection of the fixation
index and observe genetic diversity in geographic space. To generate
this map, the SPA values are taken and a table of coordinates and SPA
values is generated for each cell of the raster. The SPA values are then
extrapolated to obtain fixation index values from the linear model
obtained previously. This column is added to the coordinate table and a
raster is generated with the projected \(F_{st}\) values. Finally, a
cutoff of this raster is performed using a map of the distribution of
the species at present with a cutoff threshold of 90\% in binary format
(where \(1 =\)presence and \(0 =\) absence), since the records of the
presence localities were obtained from bases that might contain some
errors in the data (Phillips, Anderson, and Schapire 2006).

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{current-and-past-distribution}{%
\subsubsection{Current and past
distribution}\label{current-and-past-distribution}}

We show the species distribution model for \emph{Sciurus aberti} along
the time line of the projected area per time slice. We show 5 notable
moments in time line. and 5 notable points

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{all_figures/figure_3.png}
\caption{A) Species distribution model for \emph{Sciurus aberti} present
distribution. B) Relative projected area for SDM in pixels acoording a
threshold of 0.9 suitability for each past time-sliced environmental
conditions. We can observe a maximum area corresponding to the last
interglacial era and an inverted U-shape in the trend of the area along
the time considered. Also It can be noted that the actual distribution
has similar area to the coditions 120 000 years ago. B) and D)
corresponds to distribution peaks along the time scale and C) and E)
corresponds to minimum}
\end{figure}

\hypertarget{spa-and-endemic-historic-area}{%
\subsubsection{SPA and Endemic Historic
Area}\label{spa-and-endemic-historic-area}}

We get the suitability presence area (figure 2) delmited by a 0.5 and
0.9 threshold (Figure 4). The endemic region where the enviornmental
conditions 90 \% del tiempo is the Sierra Madre Occidental, in the
limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this
Area to the \emph{Sciurus aberti barbieri} subspecies. Also is notable
that at 0.5 threshold, other regions seems to have stable conditions. We
observe the area close to Coronado National Forest as well as Gila
National Forest and Mescalero Reservation as suitable areas along time
in the United States and some Areas of the Volcanic Belt in Mexico
(Puebla, north of Guerrero State and Jalisco.\\

\begin{figure}
\centering
\includegraphics{all_figures/figure_4.png}
\caption{Suitability prevalence area. Here we observe the harmonic mean
of suitability for each pixel along the distribucion from 0 to 1. The
bold lines shows the the area when the harmonic mean es above 0.5
threshold (blud line) and 0.9 threshold (black line)}
\end{figure}

\hypertarget{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
\subsubsection{Fst as a funciton of SPA and Fixation index projected in
geography}\label{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}

For all the linear regressions performed (Table 1), the only one that
had statistical significance is the one for the Sciurus aberti barberi
population with an R-squared value of 0.794.

\begin{longtable}[]{@{}ccc@{}}
\caption{R squared from linear models (fixation index as a function of
suitability prevalence area). The only R squared with statistical
significance is barberi}\tabularnewline
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
S. a. aberti Carson-SFW & 0.000 & 0.962 \\
S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
S. a. aberti MT-Zuni & 0.002 & 0.894 \\
S. a. aberti San Juan & 0.000 & 0.952 \\
S. a. barberi & 0.794 & 0.001 \\
S. a. chuscensis E & 0.002 & 0.892 \\
S. a. chuscensis W & 0.018 & 0.713 \\
S. a. ferreus Carson E & 0.040 & 0.581 \\
S. a. ferreus Pike & 0.000 & 0.967 \\
S. a. kaibabensis & 0.005 & 0.841 \\
\end{longtable}

It can be noted that S. a. aberti Coconino-Gila had an R-squared value
greater than 0 (0.131) but it is not statistically significant.
Furthermore, in our approach we imputed the value of the origin with an
Fst of zero. To show that the model hypothesis still holds, we performed
a regression for barberi without the point of origin imposed (Figure
5b), in addition to eliminating the S. aberti chuscensis populations due
to their low uncertainty in georeference, obtaining an R-squared of 0.72
(p value = 0.016).

\begin{figure}
\includegraphics[width=0.5\linewidth]{all_figures/figure_5a} \includegraphics[width=0.5\linewidth]{all_figures/figure_5b} \caption{Linear regression for S a. barberi population A) SPA in all populations B) withouth barberi and chusensis populations. We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)}\label{fig:figures-side}
\end{figure}

From the model we extrapolated the values from SPA along the geography
to new \(F_{st}\) values to get a geographical projection of fixation
index from \emph{S. a. barberi} population. (Figure 6)

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{all_figures/figure_6.png}
\caption{Projection of the \(F_{st}\) model to the geography in the
actual distribution of \emph{S aberti}}
\end{figure}

\hypertarget{references}{%
\subsubsection*{References}\label{references}}
\addcontentsline{toc}{subsubsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-beyer2020high}{}}%
Beyer, Robert M, Mario Krapp, and Andrea Manica. 2020.
{``High-Resolution Terrestrial Climate, Bioclimate and Vegetation for
the Last 120,000 Years.''} \emph{Scientific Data} 7 (1): 236.

\leavevmode\vadjust pre{\hypertarget{ref-bono2018genome}{}}%
Bono, Jeremy M, Helen K Pigage, Peter J Wettstein, Stephanie A Prosser,
and Jon C Pigage. 2018. {``Genome-Wide Markers Reveal a Complex
Evolutionary History Involving Divergence and Introgression in the
Abert's Squirrel (Sciurus Aberti) Species Group.''} \emph{BMC
Evolutionary Biology} 18 (1): 1--17.

\leavevmode\vadjust pre{\hypertarget{ref-burgin2018many}{}}%
Burgin, Connor J, Jocelyn P Colella, Philip L Kahn, and Nathan S Upham.
2018. {``How Many Species of Mammals Are There?''} \emph{Journal of
Mammalogy}. Oxford University Press US.

\leavevmode\vadjust pre{\hypertarget{ref-guisan2005predicting}{}}%
Guisan, Antoine, and Wilfried Thuiller. 2005. {``Predicting Species
Distribution: Offering More Than Simple Habitat Models.''} \emph{Ecology
Letters} 8 (9): 993--1009.

\leavevmode\vadjust pre{\hypertarget{ref-hedrick2005standardized}{}}%
Hedrick, Philip W. 2005. {``A Standardized Genetic Differentiation
Measure.''} \emph{Evolution} 59 (8): 1633--38.

\leavevmode\vadjust pre{\hypertarget{ref-karlin1968rates}{}}%
Karlin, Samuel. 1968. {``Rates of Approach to Homozygosity for Finite
Stochastic Models with Variable Population Size.''} \emph{The American
Naturalist} 102 (927): 443--55.

\leavevmode\vadjust pre{\hypertarget{ref-krapp2021statistics}{}}%
Krapp, Mario, Robert M Beyer, Stephen L Edmundson, Paul J Valdes, and
Andrea Manica. 2021. {``A Statistics-Based Reconstruction of
High-Resolution Global Terrestrial Climate for the Last 800,000
Years.''} \emph{Scientific Data} 8 (1): 228.

\leavevmode\vadjust pre{\hypertarget{ref-leonardi2023pastclim}{}}%
Leonardi, Michela, Emily Y Hallett, Robert Beyer, Mario Krapp, and
Andrea Manica. 2023. {``Pastclim 1.2: An r Package to Easily Access and
Use Paleoclimatic Reconstructions.''} \emph{Ecography}, e06481.

\leavevmode\vadjust pre{\hypertarget{ref-lira2014relationship}{}}%
Lira-Noriega, Andrés, and Joseph D Manthey. 2014. {``Relationship of
Genetic Diversity and Niche Centrality: A Survey and Analysis.''}
\emph{Evolution} 68 (4): 1082--93.

\leavevmode\vadjust pre{\hypertarget{ref-merow2013practical}{}}%
Merow, Cory, Matthew J Smith, and John A Silander Jr. 2013. {``A
Practical Guide to MaxEnt for Modeling Species' Distributions: What It
Does, and Why Inputs and Settings Matter.''} \emph{Ecography} 36 (10):
1058--69.

\leavevmode\vadjust pre{\hypertarget{ref-nei1986definition}{}}%
Nei, Masatoshi. 1986. {``Definition and Estimation of Fixation
Indices.''} \emph{Evolution} 40 (3): 643--45.

\leavevmode\vadjust pre{\hypertarget{ref-nogues2009predicting}{}}%
Nogués-Bravo, David. 2009. {``Predicting the Past Distribution of
Species Climatic Niches.''} \emph{Global Ecology and Biogeography} 18
(5): 521--31.

\leavevmode\vadjust pre{\hypertarget{ref-owens2013constraints}{}}%
Owens, Hannah L, Lindsay P Campbell, L Lynnette Dornak, Erin E Saupe,
Narayani Barve, Jorge Soberón, Kate Ingenloff, et al. 2013.
{``Constraints on Interpretation of Ecological Niche Models by Limited
Environmental Ranges on Calibration Areas.''} \emph{Ecological
Modelling} 263: 10--18.

\leavevmode\vadjust pre{\hypertarget{ref-peterson2008rethinking}{}}%
Peterson, A Townsend, Monica Papeş, and Jorge Soberón. 2008.
{``Rethinking Receiver Operating Characteristic Analysis Applications in
Ecological Niche Modeling.''} \emph{Ecological Modelling} 213 (1):
63--72.

\leavevmode\vadjust pre{\hypertarget{ref-peterson2012species}{}}%
Peterson, A Townsend, and Jorge Soberón. 2012. {``Species Distribution
Modeling and Ecological Niche Modeling: Getting the Concepts Right.''}
\emph{Natureza \& Conserva{ç}{ã}o} 10 (2): 102--7.

\leavevmode\vadjust pre{\hypertarget{ref-phillips2006maximum}{}}%
Phillips, Steven J, Robert P Anderson, and Robert E Schapire. 2006.
{``Maximum Entropy Modeling of Species Geographic Distributions.''}
\emph{Ecological Modelling} 190 (3-4): 231--59.

\leavevmode\vadjust pre{\hypertarget{ref-saupe2012variation}{}}%
Saupe, EE, V Barve, CE Myers, J Soberón, N Barve, CM Hensz, AT Peterson,
H Lc Owens, and A Lira-Noriega. 2012. {``Variation in Niche and
Distribution Model Performance: The Need for a Priori Assessment of Key
Causal Factors.''} \emph{Ecological Modelling} 237: 11--22.

\leavevmode\vadjust pre{\hypertarget{ref-singarayer2010high}{}}%
Singarayer, Joy S, and Paul J Valdes. 2010. {``High-Latitude Climate
Sensitivity to Ice-Sheet Forcing over the Last 120 Kyr.''}
\emph{Quaternary Science Reviews} 29 (1-2): 43--55.

\leavevmode\vadjust pre{\hypertarget{ref-soberon2005interpretation}{}}%
Soberon, Jorge, and A Townsend Peterson. 2005. {``Interpretation of
Models of Fundamental Ecological Niches and Species' Distributional
Areas.''}

\leavevmode\vadjust pre{\hypertarget{ref-soberon2009niches}{}}%
Soberón, Jorge, and Miguel Nakamura. 2009. {``Niches and Distributional
Areas: Concepts, Methods, and Assumptions.''} \emph{Proceedings of the
National Academy of Sciences} 106 (supplement\_2): 19644--50.

\leavevmode\vadjust pre{\hypertarget{ref-soberon2020shape}{}}%
Soberón, Jorge, and A Townsend Peterson. 2020. {``What Is the Shape of
the Fundamental Grinnellian Niche?''} \emph{Theoretical Ecology} 13 (1):
105--15.

\leavevmode\vadjust pre{\hypertarget{ref-thorup2021response}{}}%
Thorup, Kasper, Lykke Pedersen, Rute R Da Fonseca, Babak Naimi, David
Nogués-Bravo, Mario Krapp, Andrea Manica, et al. 2021. {``Response of an
Afro-Palearctic Bird Migrant to Glaciation Cycles.''} \emph{Proceedings
of the National Academy of Sciences} 118 (52): e2023836118.

\leavevmode\vadjust pre{\hypertarget{ref-upham2019inferring}{}}%
Upham, Nathan S, Jacob A Esselstyn, and Walter Jetz. 2019. {``Inferring
the Mammal Tree: Species-Level Sets of Phylogenies for Questions in
Ecology, Evolution, and Conservation.''} \emph{PLoS Biology} 17 (12):
e3000494.

\leavevmode\vadjust pre{\hypertarget{ref-valdes2017bridge}{}}%
Valdes, Paul J, Edward Armstrong, Marcus PS Badger, Catherine D
Bradshaw, Fran Bragg, Michel Crucifix, Taraka Davies-Barnard, et al.
2017. {``The BRIDGE HadCM3 Family of Climate Models: HadCM3@ Bristol V1.
0.''} \emph{Geoscientific Model Development} 10 (10): 3715--43.

\leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%
Wright, Sewall. 1943. {``Isolation by Distance.''} \emph{Genetics} 28
(2): 114.

\end{CSLReferences}

\end{document}

### 2023-07-08 manuscript_english_word.Rmd
adding preable.tex

---
title: "Suitability prevalence area index in late quaternary explains genetic"
author: "Norma Hernandez, Angel Robles, Nathan Upham"
date: "2023-07-08"
output: word_document
bibliography: R21_citation.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract
As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].

Our results reveal that fitness prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

## Introduction

The current distributions of species do not always correspond to their historical distributions over evolutionarily significant periods. This is because environmental conditions are not static over time; species tend to distribute where conditions are most favorable.

It follows from this relationship between environment and species distribution that population size also varies with time. This variation in population size is related to the effective population size $N_e$ [@karlin1968rates], so that an index reflecting changes in environmental conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index $F_{st}$. Thus, it is possible to relate patterns of changes in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.

With this approach we can predict the geographic pattern of population structure from environmental information. This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting] and population genetics theories [@lira2014relationship].

In this work we propose a method to find the Suitability Prevalence Area (SPA) as an index with a double purpose: 1) to find endemic areas to delimit the historical distribution of the species and 2) to explain the patterns of genetic diversity.

To obtain the SPA, we performed a historical reconstruction of the geographical range back to 120 000 BC at 2 000 year intervals and recorded the environmental suitability at each site.

Subsequently, we delimited historical endemic areas to locations where the prevalence of suitability remained at 90% during this period. From the fixation index calculated with respect to populations in the historical endemic areas, a statistical model of the fixation index as a function of SPA was performed. With this statistical model, the fixation index values were projected to the current distribution to obtain a map with the geographic pattern of this index.

### Theoretical basis

#### Environment, fitness and suitability

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.

That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates. However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.

Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$. Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space. Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.

![Hypothetical niche in environmental space. An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}

This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of fitness and environment. In that study they tested the validity with monthly variables and empirical values on a short time scale. relating life history parameters and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$. In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly. in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Es decir:

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$

#### Genetic variation and effective population size

On the other hand, loss of heterozygosity may be related to the amount of genetic variation present in the absence of mutation and selection, so theoretically, one would expect a strong correlation between effective population size and heterozygosity on the basis of population genetics. We start from the assumption that in the absence of gene flow between subpopulations, the rate of fixation by isolation depends on the effective population size $N_e$ when subpopulations diverge from a common ancestor in $t$ generations. [@hedrick2005standardized: @wright1943isolation]:

$$
F_{st}(t) = 1 - e^{-kt/N_e } 
$$

where we assume, according to Nei [@nei1986definition], that genetic drift within each subpopulation causes the average heterozygosity between populations ($H_s$) to approach zero ($F_{st} = 1 - H_s/H_t$), i.e., for $t=0$ $F_{st} = 0$ and for very large $t$ the index is $1$.
that is, for $t=0$ $F_{st} = 0$ and for very large $t$ the index is $1$. We have included a constant $k$ without loss of generalization.

Using $e(x) = \sum_{k=0}^{\infty} \frac{x^k}{k!} = 1 + x + \frac{x^2}{2} + ...$ and leaving to first order for $N_e$ large enough

$$
F_{st} \approx \frac{kt}{N_e} + C
$$

Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)$. We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close.

$$
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
$$

#### Relationship between the fixation index and environmental suitability
We can infer locally for each population $N_i$ in the geography expect to be proportional to the environmental suitability for each time considered, 
For simplification in the index we select a $N_i$ population in the geography and infer that locally, this population is proportional to the environmental suitability in each time considered, i. e. $N_{j} \propto S_{j}(\vec{e}_{j})$  (each index $j$ is for the generation time), as we describe above, so we can asume an harmonic mean of suitabilities $\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j}  = \frac{1}{S}$ and substitue in the previous equation:

$$
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{S_j(\vec{e_j})}
$$

We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time. Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the fitness prevalence for an $i$ site in the geography:

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic mean of fitness *Suitability prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.

## Methods

We summarize our 5 steps predictive approach in the figure 2.

<p align="center">
![Realized niche in environmental space. A point dynamically moves along different time slices](all_figures/figure_2.png){width=30%}
</p>

We propose a method to find the Suitability Prevalence Area (SPA) as an index with a dual purpose: 1) to find endemic areas to delimit the historical distribution of species and 2) to explain patterns of genetic diversity.

The method for estimating SPA consists of a series of 4 steps:

1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
4. Generate a harmonic mean suitability map for all time slices.

Subsequently, one application of the SPA is to explain genetic diversity (the fixation index) as a function of the SPA and project it to the geography. For this it is necessary to perform a series of steps, which consist of:

- Obtain georeferenced points of genetically structured populations with fixation index values.
- Extract values from the SPA map with these points. 
- A linear model of the fixation index value with respect to the SPA is performed.
- With the linear model all SPA values of suitability are interpolated.
- A map is generated with all the values interpolated by the model.

To test our model we use the Aberti squirrel (*Sciurus aberti*) as a case study because:

- Since it is a medium-sized mammal (0.5 kg) its identification is easy.
- The subspecies are geographically isolated, since they are found in pine forests surrounded by desert. Therefore we can consider the assumption of isolation by distance [@wright1943isolation]. 
- There are sufficient occurrence records for the entire range of the known distribution.
- It has low speciation rate (0.25) [@upham2019inferring], therefore, we assume that its fundamental niche is stable over time.
- There is fixation index information [@bono2018genome] for different populations of the subspecies, which is useful for using SPA as a way to explain genetic differentiation between populations.

### Unique occurrence records 
We have downloaded geographic information on Sciurus aberti from the open access platform Global Biodiversity Information Facility (GBIF). Subsequently, we eliminated duplicate records, those without precise coordinates and those with uncorroborated collection information. Once the unique occurrence records were obtained, we filtered the geographic space in order to avoid overpredictions in the spatial distribution models caused by the agglomeration of occurrence points. The filtering consisted of superimposing the records in the geographic space, then we made 5 latitudinal slices of 5 degrees of separation between each one, from latitude 20 degrees south to latitude 45 degrees north, taking as criteria the known distribution of the species and the ecoregions of North America, level 1 (downloaded from the US Environmental Protection Agency, 2010). Once the geographic space was segmented, we left the same number of records for each band (30 records per band) obtaining a total of 150 unique records of presence.

### Environmental characterization
The environmental information to generate the models, from the present and the reconstruction to the past, was obtained from Pastclim 1.2, an R statistical software package designed to download and manipulate paleoclimatic datasets. In our particular case we chose the Beyer at al. @beyer2020high set as it has environmental information available up to 120 000 BE, significantly evolutionary time, at 2000 year time intervals based on Global Circulation Models HadCM3 [@singarayer2010high] and HadAM3H [@valdes2017bridge].  The environmental dataset has 17 bias-corrected bioclimatic variables, reduced spatial scale and spatial resolution of 0.5°square cells.

### Accessibility region ($\textbf{M}$).
Species distribution models according to the BAM diagram [@soberon2005interpretation] states that the geographic range occupied by a species ($\textbf{G}_o$) is the region of appropriate assemblages in terms of abiotic conditions. The region $\textbf{M}$ represents in geography areas in which the species has access due to its colonization capabilities and the structure of geographic barriers within a specific period of time [@soberon2009niches].

In this study we selected a set of North American ecoregions for modeling to delimit our Accessibility region. We took into account ecoregions that overlap with the known distribution of the species and also took into account neighboring ecoregions to increase the potential accessibility in projection scenarios to the past. Thus, regions 5, 6, 9, 10, 12, 13 of the level I ecoregions (<https://www.epa.gov/>) were selected.

### Niche models 
#### Current distribution
Estimating niche boundaries for species occurrence is called ecological niche modeling (ENM), and when the emphasis is on geographic distribution, it is known as species distribution modeling (SDM) [@guisan2005predicting; @peterson2012species; @saupe2012variation]. We model the current distribution of S. aberti using the MAXENT 3.4.4 algorithm [@phillips2006maximum], which is a correlative model based on the maximum entropy principle used to estimate species distributions. The output of Maxent is a relative occurrence rate (ROR) interpreted as a probability of habitat suitability given observed environmental conditions by presence-only points [@merow2013practical].

The parameterization was performed with the default values of the program with the exception of the "extrapolate" and "clamping" options to avoid artificial extrapolations in the extreme values of the climatic variables used in the models. To calibrate the models we used 70% of the records and the remaining 30% to validate them.

In order to give statistical certainty and to consider uncertainty, as well as to reduce overfitting and to have a suitable model to extrapolate to past scenarios, 10 replications were performed for each model using a cross-validation with the training data. In the end, the average model of these replications was considered as the result.

We evaluated the statistical significance of the resulting models for the present using the partial ROC test [@peterson2008rethinking], which is a modification of the ROC (receiver Operation Characteristic) test. The results of this test are proportions (ratios) of the area under the curve of the model with respect to a null model, product of repetitions that allow to statistically evaluate the areas under the curve (AUC) in relation to that expected by chance [@peterson2008rethinking], where a value derived by chance would be $0$ and an acceptable value, according to the proportion of minimum omission errors tolerated in the model, would be greater than $1$. A different set of test data was used on the average model of the replicates and in this way the reduction of overfitting was guaranteed.

#### Reconstruction to the past
The average model obtained from Maxent was extrapolated to environmental conditions into the past. Environmental layers were taken from pastclim [@leonardi2023pastclim] in 2000 year time slices to cover a total of 120 000 years. For each scenario into the past, 10 replicates were cross-validated and averaged.

#### Analogy of environmental conditions 
To be certain that areas in the past are climatically analogous to the present we implemented the mobility-orient parity (MOP) test. This method generates a value from $0$ to $1$ according to the environmental analogy of past conditions to the present to eliminate geographic areas in the projections to the past outside of climates analogous to the present [@owens2013constraints].

### Suitability Prevalence Area (SPA)
Finally, to generate the SPA, both the average projection of the present and all the average projections of the past are considered. The output files are read in raster format.

A harmonic mean is performed as follows:
For each layer the inverse of each cell value is obtained. Then the inverse values are summed by cells and divided by the number of scenarios. Finally, the inverse of each cell is obtained. It is important to consider that the Maxent outputs may contain very small values close to zero in many cells, for this reason we added a small amount of $0.001$ to all the values in all the maps so that when obtaining the harmonic mean we could have an inverse value for values close to zero without altering the biological sense of environmental suitability.

### Endemic historic Area
To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $mathbf{S}$ equation. This threshold is the proportion of time that pixel has had ideal conditions. For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.

### Fixation index as a functino of SPA

To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered. For this work we considerer the data from @@bono2018genome. It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form. The matrix was transformed into a table as pairs between populations with index values between each pair. The values of the population with itself, i.e. $0$, were included. Subsequently, georeferences were obtained for each population considered in the fixation index matrix. From this geo-referencing the SPA value for each population was extracted. Finally, the SPA value was aligned in the table of fixation index pairs. In total the table has 4 columns: source population, target population, fixation index and SPA of the target population. This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$. An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.

### Fixation index projected in geography
The linear model that presents the highest value of $R^2$ is considered suitable to make a geographic projection of the fixation index and observe genetic diversity in geographic space.  To generate this map, the SPA values are taken and a table of coordinates and SPA values is generated for each cell of the raster. The SPA values are then extrapolated to obtain fixation index values from the linear model obtained previously. This column is added to the coordinate table and a raster is generated with the projected $F_{st}$ values. Finally, a cutoff of this raster is performed using a map of the distribution of the species at present with a cutoff threshold of 90% in binary format (where $1 =$presence and $0 =$ absence), since the records of the presence localities were obtained from bases that might contain some errors in the data [@phillips2006maximum].

## Results

### Current and past distribution
We show the species distribution model for *Sciurus aberti*  along the time line of the projected area per time slice. We show 5 notable moments in time line. 
and 5 notable points 
<p align="center">
![A) Species distribution model for *Sciurus aberti* present distribution. B) Relative projected area for SDM in pixels acoording a threshold of 0.9 suitability for each past time-sliced environmental conditions. We can observe a maximum area corresponding to the last interglacial era and an inverted U-shape in the trend of the area along the time considered. Also It can be noted that the actual distribution has similar area to the coditions 120 000 years ago. B) and D) corresponds to distribution peaks along the time scale and C) and E) corresponds to minimum  ](all_figures/figure_3.png){width=100%}
</p>

### SPA and Endemic Historic Area
We get the suitability presence area (figure 2) delmited by a 0.5 and 0.9 threshold (Figure 4). The endemic region where the enviornmental conditions 90 \% del tiempo is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this Area to the *Sciurus aberti barbieri* subspecies. Also
is notable that at 0.5 threshold, other regions seems to have stable conditions. We observe the area close to Coronado National Forest as well as Gila National Forest and Mescalero Reservation as suitable areas along time in the United States and some Areas of the Volcanic Belt in Mexico (Puebla, north of Guerrero State and Jalisco.  
<p align="center">
![Suitability prevalence area. Here we observe the harmonic mean of suitability for each pixel along the distribucion from 0 to 1. The bold lines shows the the area when the harmonic mean es above 0.5 threshold (blud line) and 0.9 threshold (black line)](all_figures/figure_4.png){}
</p>

### Fst as a funciton of SPA and Fixation index projected in geography

For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an R-squared value of 0.794.

```{r, echo=FALSE}
aberti_pops <- read.csv(file = "all_tables/fst_SPA_rsquared.csv")

knitr::kable(aberti_pops, "simple", align = "ccc", caption = "R squared from linear models (fixation index as a function of suitability prevalence area). The only R squared with statistical significance is barberi")
```

It can be noted that S. a. aberti Coconino-Gila had an R-squared value greater than 0 (0.131) but it is not statistically significant. Furthermore, in our approach we imputed the value of the origin with an Fst of zero. To show that the model hypothesis still holds, we performed a regression for barberi without the point of origin imposed (Figure 5b), in addition to eliminating the S. aberti chuscensis populations due to their low uncertainty in georeference, obtaining an R-squared of 0.72 (p value = 0.016).

```{r, figures-side, fig.show="hold", out.width="50%",  fig.cap="Linear regression for S a. barberi population A) SPA in all populations B) withouth barberi and chusensis populations. We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)", echo=FALSE}
knitr::include_graphics(c("all_figures/figure_5a.png","all_figures/figure_5b.png"))
```

From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S. a. barberi* population. (Figure 6)

<p align="center">
![Projection of the $F_{st}$ model to the geography in the actual distribution of *S aberti*](all_figures/figure_6.png){width=80%}
</p>

## Discussion

### References

### 2023-07-08 manuscript_english.Rmd
Adding word document

--- 2023-07-08 manuscript_english_word.Rmd
@@
 ---
-title: "Suitability prevalence area index in late quaternary explains genetic"
-author: "Norma Hernandez, Angel Robles, Nathan Upham"
-date: "2023-07-08"
-output: word_document
+title: Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels
+author: "Norma Hernandez, Angel Robles Nathan Upham"
+date: "2023-06-27"
+includes:
+  in_header: preamble.tex
+output: 
+  pdf_document:
+    fig_caption: yes
+    latex_engine: xelatex 
+    keep_tex: true
 bibliography: R21_citation.bib
 ---
+
 ```{r setup, include=FALSE}
@@
 
-## Discussion
-
 ### References

### 2023-07-10 manuscript_english.Rmd
upgrade manuscirpt english:

@@
 
-We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables.
-Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
-that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.
-
-That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
-If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
-
+For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$.
+We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$.
+Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions.
+If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$.
+i.
+e.
+the set of environmental conditions for which values of fitness are capable of sustaining populations.
+We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$.
 This is a way of expressing the fundamental niche as a function of environmental conditions.
 So we can map each point of geography to ambient space.
-Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
-However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates.
-However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.
-
+and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
+However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$.
 Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.
-Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
-$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.
-
-We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space.
-Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.
-
-![Hypothetical niche in environmental space.
-An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
 
@@
 This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$.
-In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.
+That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$.
+In a practical way, instead of $\mathbf{N}_F$, we approximated
+$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
+(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.
 
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly.
-in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
+The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
 
-The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
-
-Es decir:
-
+Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
 $$
@@
 $$
-Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
+
+![Hypothetical niche in environmental space.
+An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
+
+Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates.
+So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:
+
+$$
+N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
+$$
 
@@
 
-1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
+1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
 2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
-3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
-4. Generate a harmonic mean suitability map for all time slices.
+3. Characterize past environments and retain projections only for sites with analogous environmental conditions to the present.
+4. Calculate the harmonic mean of the suitability for all the sites all time slices.
 
@@
 ### Endemic historic Area
-To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $mathbf{S}$ equation.
+To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $\mathbf{S}$ equation.
 This threshold is the proportion of time that pixel has had ideal conditions.

### 2023-07-10 manuscript_english.tex
upgrade manuscirpt english:

@@
 
-We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
+For a given region \(\mathbf{G}\) in te geography, at any point in time,
+we have a set of existing environmental conditions
+\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
+We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
 environmental variables, that is, a point in the space of \(v\)
-environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
-relates fitness to environment, that is, for a species with
-non-overlapping generations, the population's net reproductive rate
-\(R_i\) is a measure of fitness that depends on environmental
-conditions.
-
-That is, the sites in the geography \(\mathbf{G}_o\) that present
-positive growth rates are related to favorable conditions. If we take
-the set of places in ambient space where these values are greater than a
-threshold, we can define a volume in multidimensional space such that
-\(R(\vec{e})\) is always positive and greater than this threshold, that
-is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
-
-This is a way of expressing the fundamental niche as a function of
-environmental conditions. So we can map each point of geography to
-ambient space. Thus, for each site in the geography there will
-correspond a site in the environmental space with different
-environmental conditions and different growth rates. However all sites
-within the region defined by \(\mathbf{N}_F\) will have viable
-populations and positive rates. However, we assume that the environment
-is dynamic and each geographic point is static, so for each site in
-geographic space this growth rate depends on the environment and
-therefore is different at each time.
-
-Thus locally the growth rate will be related to the local population
-size as long as it is within the boundary defined by \(\mathbf{N}_F\).
-Therefore, the local population size as a function of the intrinsic rate
-that depends on the change in environmental conditions over time is:
-\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
-population size is a function of ideal environmental conditions.
-
-We start by considering this defined volume of the set of \(R_0\) as a
-set of trajectories for each point from geographic space to
-environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
-the population grows, that is,
-\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
-any local population will have a positive growth rate.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
-\caption{Hypothetical niche in environmental space. An arbitrary point
-moves within the ambient space entering and exiting the volume (niche)
-where the intrinsic growth rate is positive.}
-\end{figure}
+environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
+\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
+that is, for a species with non-overlapping generations, the
+population's net reproductive rate \(R_0\) is a measure of fitness that
+depends on environmental conditions. If we take the set of places in
+environmental space where these values are greater than a threshold, we
+can define a volume in multidimensional space such that \(R(\vec{e})\)
+is always positive and greater than this threshold, that is
+\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
+environmental conditions for which values of fitness are capable of
+sustaining populations. We call
+\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
+expressing the fundamental niche as a function of environmental
+conditions. So we can map each point of geography to ambient space. and,
+for each site in the geography there will correspond a site in the
+environmental space with different environmental conditions and
+different growth rates. However all sites within the region defined by
+\(\mathbf{N}_F\) will have viable populations and positive rates
+depending on time because we assume that the environment is dynamic and
+each geographic point is static, so for each site in geographic space
+this growth rate depends on the environment and therefore is different
+at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
+will be related to the local population size as long as it is within the
+boundary defined by \(\mathbf{N}_F\).
 
@@
 That is, the fitness-related environmental suitability can be any
-function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
-\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
-of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
-approximated from points of presence in the geography obtained from
-databases, numerically approximating a convex surface with a surface
-\(\mathbf{N}_R\) of the environmental space.
+function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
+\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
+values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
+practical way, instead of \(\mathbf{N}_F\), we approximated
+\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
+existing niche) from known presence points in the geography
+\(\mathbf{G}\) obtained from emprical observations stored in database,
+and map to \(\mathbf{E}\) in order to numerically fitting a convex
+volume of the environmental space that enclose the known information.
+
+The objective of our approach is not to know the function of \(R_0\) but
+rather to relate this local variation and the function \(\mathbf{N}_F\)
+with a function \$ S\_i(~vec\{e\}\_i)\$ of environmental suitability
+that can be estimated from known methods of applied mathematics and
+machine learning (for example the case of Maxent, Phillips, Anderson,
+and Schapire (2006)).
+
+Population size locally is a function of environmental suitability,
+where if suitability is \(0\) the population is \(0\) and if suitability
+is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) \[
+N_i \propto  S_i(\vec{e}_i)
+\]
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
+\caption{Hypothetical niche in environmental space. An arbitrary point
+moves within the ambient space entering and exiting the volume (niche)
+where the intrinsic growth rate is positive.}
+\end{figure}
 
@@
 point in the geography is static and has a trajectory in the
-environmental space and that for each trajectory the fitness function
-changes and therefore the suitability function changes but not
-accordingly. in a homogeneous way, if not more so that each site
-presents different displacement rates and different demographic rates.
-
-The objective of this theoretical approach is not to know the shape of
-the function of \(R_0\) but rather to find this local variation relating
-this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
-\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
-estimated from known methods of applied mathematics and machine learning
-(for example the case of Maxent, Phillips, Anderson, and Schapire
-(2006)).
-
-Es decir:
+environmental space and for each trajectory the fitness function changes
+and therefore the suitability function changes but not in a homogeneous
+way, so each site presents different displacement rates in envrionmental
+spaces and as a consequence, different demographic rates. So we can
+establish that for each site \(i\) in geographic space \(\mathbf{G}\),
+there is a population size \(N_i\) that changes with respect to a
+fitness function dependent on time-varying environmental conditions:
 
 \[
-N_i \propto  S_i(\vec{e}_i)
-\] Population size locally is a function of environmental suitability,
-where if suitability is \(0\) the population is \(0\) and if suitability
-is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)
+N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
+\]
 
@@
 threshold is applied to generate a binary map according to the
-\(mathbf{S}\) equation. This threshold is the proportion of time that
+\(\mathbf{S}\) equation. This threshold is the proportion of time that
 pixel has had ideal conditions. For our study, we consider the value of

### 2023-07-10 manuscript_english.tex
Updates from Overleaf

--- 2023-07-08 manuscript_english.tex
@@
 
-\title{Suitability prevalence area index in late quaternary explains
-genetic diversity in Tassel eared Squirrels}
+\title{Suitability prevalence area index in late Quaternary explains
+genetic diversity in Tassel-eared Squirrels}
 \author{Norma Hernandez, Angel Robles Nathan Upham}
@@
 
-As a case study we consider squirrels (\emph{Sciurus aberti}), a species
-currently distributed in disjunct patches from the southern Rocky
+As a case study we consider Tassel-earred Squirrels (\emph{Sciurus aberti}), a species
+currently distributed in disjunct patches of pine forest from the southern Rocky
 Mountains in the United States to the northern Sierra Madre Occidental
-in Mexico for which it is possible to find reliable information on both
-the genetic structure of its populations and its current distribution
-(Bono et al. 2018; Burgin et al. 2018).
-
-Our results reveal that fitness prevalence corresponds to the fixation
-index of \emph{S. aberti} populations with respect to a source
-population. Populations closer to the historical endemic area present a
-higher genetic diversity and a lower \(F_{st}\) value. Finally, this
-study allows us to add a biogeographic explanation to the results
-obtained with population genetic methods and to generate maps of this
-structure as tools to support conservation with a perspective that
-integrates both population genetics and historical patterns of species
-distribution.
+in Mexico for which reliable population-genetic and occurrence information is available. Our results reveal that the genetic fixation
+index (\(F_{st}\)) of \emph{S. aberti} populations with respect to a source
+population correspond to the area of greatest prevalence through time, as modeled over the last 120,000 years. Populations closer to the historical endemic area display a
+higher genetic diversity and a lower \(F_{st}\) value.
+This result suggests a general biogeographic explanation for the common observation of geographic clines of genetic diversity within species.
+By generating a time series of habitat suitability maps for a given species,  the suitability prevalence area can be estimated.
+Thus, integrating  population-genetic data with estimates of historical patterns of species
+distribution holds promise for improving conservation and management efforts.
 
@@
 
-The current distributions of species do not always correspond to their
-historical distributions over evolutionarily significant periods. This
-is because environmental conditions are not static over time; species
-tend to distribute where conditions are most favorable.
+The current distributions of species may not correspond to their
+historical distributions over evolutionary significant time periods. This
+is because environmental conditions are not static over time, and the populations of a species
+tend to 'track' environments, distributing where conditions are most favorable in the absence of barriers (REF).
 
@@
 
-We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
-environmental variables, that is, a point in the space of \(v\)
+We define \(\vec{e}_i =(e_1, e_2, ..., e_n )\) as a vector of
+environmental variables, that is, a point in the space of \(n\)
 environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
@@
 \(R(\vec{e})\) is always positive and greater than this threshold, that
-is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
+is \(\mathbf{N}_F = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
 
@@
 within the region defined by \(\mathbf{N}_F\) will have viable
-populations and positive rates. However, we assume that the environment
+populations and positive rates. We assume that the environment
 is dynamic and each geographic point is static, so for each site in
@@
 that depends on the change in environmental conditions over time is:
-\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
+\(N_i(t+1) = R_0(\vec{e}_i(t) ) N_i(t)\). That is, the locally effective
 population size is a function of ideal environmental conditions.
@@
 set of trajectories for each point from geographic space to
-environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
+environmental space. Considering this rate fixed \(R_0\) when
 the population grows, that is,
@@
 
-This equation was proposed by Soberon and Peterson Soberón and Peterson
+This equation was proposed by Soberon and Peterson
 (2020) to relate demographics to a function of fitness and environment.
@@
 relate this growth rate to an environmental suitability function bounded
-between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
+between \(0\) and \(1\), for example, for a unimodal multivariate probability density function,
 it would be equivalent to the distance from the maximum of the function.
@@
 this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
-\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
+\(S_i(\vec\{e\}\_i)\) of environmental suitability that can be
 estimated from known methods of applied mathematics and machine learning
 (for example the case of Maxent, Phillips, Anderson, and Schapire
-(2006)).
-
-Es decir:
+(2006)):
 

### 2023-07-11 manuscript_english.Rmd
change abstract

@@
 ## Abstract
-As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
-
-Our results reveal that fitness prevalence corresponds to the fixation index of *S.
-aberti* populations with respect to a source population.
-Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value.
-Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
-
-## Introduction
 
@@
 
+As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
+
+Our results reveal that suitability prevalence corresponds to the fixation index of *S.
+aberti* populations with respect to a source population.
+Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value.
+Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
+
+## Introduction
+
 ### Theoretical basis
@@
 This is a way of expressing the fundamental niche as a function of environmental conditions.
-So we can map each point of geography to ambient space.
-and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
+So we can correspong each point of geography to environmental space,  with different environmental conditions and different growth rates.
 However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$.
@@
 
-The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
+The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $S_i(\vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
 
@@
 ![Hypothetical niche in environmental space.
-An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
+An arbitrary point moves within the environmental space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
 
@@
 Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ...
-+ \frac{1}{N_\tau} } \Big)$.
-We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close.
++ \frac{1}{N_\tau} } \Big)$ for each $i$ site in $\mathbf{G}$.
+We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close:
 
 $$
-F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
+F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
 $$
@@
 We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time.
-Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the fitness prevalence for an $i$ site in the geography:
+Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the suitability prevalence for an $i$ site in the geography:
 
@@
 $$
-If we consider all sites in the available geography, we call this harmonic mean of fitness *Suitability prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
+If we consider all sites in the available geography, we call this harmonic  *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
 environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred.
@@
 The parameterization was performed with the default values of the program with the exception of the "extrapolate" and "clamping" options to avoid artificial extrapolations in the extreme values of the climatic variables used in the models.
-To calibrate the models we used 70% of the records and the remaining 30% to validate them.
+To calibrate the models we used 70\% of the records and the remaining 30\% to validate them.
 
@@
 For this work we considerer the data from @@bono2018genome.
-It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form.
+It consideres 10 populations along the known distribution and provides values of $F_{st}$ between this populations in matrix form.
 The matrix was transformed into a table as pairs between populations with index values between each pair.

### 2023-07-11 manuscript_english.tex
change abstract

--- 2023-07-10 manuscript_english.tex
@@
 rather to relate this local variation and the function \(\mathbf{N}_F\)
-with a function \$ S\_i(~vec\{e\}\_i)\$ of environmental suitability
-that can be estimated from known methods of applied mathematics and
-machine learning (for example the case of Maxent, Phillips, Anderson,
-and Schapire (2006)).
+with a function \(S_i(\vec{e}_i)\) of environmental suitability that can
+be estimated from known methods of applied mathematics and machine
+learning (for example the case of Maxent, Phillips, Anderson, and
+Schapire (2006)).
 
@@
 \item
-  model the potential distribution of a taxon with present-day
+  Model the potential distribution of a taxon with present-day
   environmental conditions and obtain an environmental suitability map.
@@
 \item
-  Characterize past analogous climates and retain projections only for
-  sites with climates analogous to the present.
+  Characterize past environments and retain projections only for sites
+  with analogous environmental conditions to the present.
 \item
-  Generate a harmonic mean suitability map for all time slices.
+  Calculate the harmonic mean of the suitability for all the sites all
+  time slices.
 \end{enumerate}

### 2023-07-11 manuscript_english_word.Rmd
change abstract

@@
 ---
-title: "Suitability prevalence area index in late quaternary explains genetic"
+title: "Suitability prevalence area index in late quaternary explains genetic diversity"
 author: "Norma Hernandez, Angel Robles, Nathan Upham"
@@
 
-Our results reveal that fitness prevalence corresponds to the fixation index of *S.
+Our results reveal that suitability prevalence corresponds to the fixation index of *S.
 aberti* populations with respect to a source population.
@@
 
-We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables.
-Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
-that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.
-
-That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
-If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
-
+For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$.
+We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$.
+Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions.
+If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$.
+i.
+e.
+the set of environmental conditions for which values of fitness are capable of sustaining populations.
+We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$.
 This is a way of expressing the fundamental niche as a function of environmental conditions.
-So we can map each point of geography to ambient space.
-Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
-However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates.
-However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.
-
+So we can correspong each point of geography to environmental space,  with different environmental conditions and different growth rates.
+However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$.
 Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.
-Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
-$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.
-
-We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space.
-Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.
-
-![Hypothetical niche in environmental space.
-An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
 
@@
 This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$.
-In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.
+That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$.
+In a practical way, instead of $\mathbf{N}_F$, we approximated
+$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
+(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.
 
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly.
-in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
+The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $S_i(\vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
 
-The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
-
-Es decir:
-
+Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
 $$
@@
 $$
-Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
+
+![Hypothetical niche in environmental space.
+An arbitrary point moves within the environmental space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
+
+Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates.
+So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:
+
+$$
+N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
+$$
 

### 2023-07-11 manuscript_english.tex
update tex

@@
 
-As a case study we consider squirrels (\emph{Sciurus aberti}), a species
-currently distributed in disjunct patches from the southern Rocky
-Mountains in the United States to the northern Sierra Madre Occidental
-in Mexico for which it is possible to find reliable information on both
-the genetic structure of its populations and its current distribution
-(Bono et al. 2018; Burgin et al. 2018).
-
-Our results reveal that fitness prevalence corresponds to the fixation
-index of \emph{S. aberti} populations with respect to a source
-population. Populations closer to the historical endemic area present a
-higher genetic diversity and a lower \(F_{st}\) value. Finally, this
-study allows us to add a biogeographic explanation to the results
-obtained with population genetic methods and to generate maps of this
-structure as tools to support conservation with a perspective that
-integrates both population genetics and historical patterns of species
-distribution.
-
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
 The current distributions of species do not always correspond to their
@@
 
+As a case study we consider squirrels (\emph{Sciurus aberti}), a species
+currently distributed in disjunct patches from the southern Rocky
+Mountains in the United States to the northern Sierra Madre Occidental
+in Mexico for which it is possible to find reliable information on both
+the genetic structure of its populations and its current distribution
+(Bono et al. 2018; Burgin et al. 2018).
+
+Our results reveal that suitability prevalence corresponds to the
+fixation index of \emph{S. aberti} populations with respect to a source
+population. Populations closer to the historical endemic area present a
+higher genetic diversity and a lower \(F_{st}\) value. Finally, this
+study allows us to add a biogeographic explanation to the results
+obtained with population genetic methods and to generate maps of this
+structure as tools to support conservation with a perspective that
+integrates both population genetics and historical patterns of species
+distribution.
+
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
+
 \hypertarget{theoretical-basis}{%
@@
 expressing the fundamental niche as a function of environmental
-conditions. So we can map each point of geography to ambient space. and,
-for each site in the geography there will correspond a site in the
-environmental space with different environmental conditions and
+conditions. So we can correspong each point of geography to
+environmental space, with different environmental conditions and
 different growth rates. However all sites within the region defined by
@@
 \caption{Hypothetical niche in environmental space. An arbitrary point
-moves within the ambient space entering and exiting the volume (niche)
-where the intrinsic growth rate is positive.}
+moves within the environmental space entering and exiting the volume
+(niche) where the intrinsic growth rate is positive.}
 \end{figure}
@@
 \(\tau\) generations
-\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
-We substitute the harmonic mean into the above equation, considering
-that although the number of generations is not exactly equal to the time
-of the fixation index, they are expected to be very close.
+\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\)
+for each \(i\) site in \(\mathbf{G}\). We substitute the harmonic mean
+into the above equation, considering that although the number of
+generations is not exactly equal to the time of the fixation index, they
+are expected to be very close:
 
 \[
-F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
+F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
 \]
@@
 loss of heterozygosity (through the fixation index \(F_{st}\)) is
-inversely proportional to the fitness prevalence for an \(i\) site in
-the geography:
+inversely proportional to the suitability prevalence for an \(i\) site
+in the geography:
 
@@
 \] If we consider all sites in the available geography, we call this
-harmonic mean of fitness \emph{Suitability prevalence Area}
+harmonic \emph{Suitability Prevalence Area}
 \(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
@@
 (2018). It consideres 10 populations along the known distribution and
-provides values of \(F_st\) between this populations in matrix form. The
-matrix was transformed into a table as pairs between populations with
-index values between each pair. The values of the population with
+provides values of \(F_{st}\) between this populations in matrix form.
+The matrix was transformed into a table as pairs between populations
+with index values between each pair. The values of the population with
 itself, i.e.~\(0\), were included. Subsequently, georeferences were

### 2023-07-11 manuscript_english.Rmd
expand introduction

@@
 ## Introduction
+
+Relating the genetic structure of the populations of a species to its geographic distribution is a complicated challenge from the point of view of biogeography.
+The task is made easier if evolutionary history is incorporated on both sides of the relationship.
+On the one hand, giving depth and an evolutionary perspective to the geographic distribution of a taxon relies primarily on ecological niche theory, and is driven by the massive amount of information currently available from both digitized biological collections and citizen science (through which records of taxa occurrence are
+through which records of taxa occurrence are available) as well as from environmental information available for geographic information systems and climate models.
+information systems and accurate climatological models of both the present and the past.
+
+This change in reproductive success is driven by a set of factors, however we propose in this work a conceptual framework where we consider only environmental factors at a macroecological scale, inspired by the concept of Grinnellian niches (those composed by sets of points of non-interactive variables in multidimensional spaces), seeking as a principle the necessary conditions for fitness and not the response of taxa.
+points of non-interactive variables in multidimensional spaces), seeking as a principle the conditions necessary for fitness and not the response of taxa.
+
+That is, we simplify our model (as is done in the physical sciences) to link through fitness the changes in biological populations and their structure, observed through population genetics, with patterns of spatial distribution over time.
+Thus, the goal is to link biogeographic patterns and evolutionary processes of populations in a single conceptual framework.
+
+It is from this Grinellian niche where we start, assuming that taxa in general have to be distributed in the geography where environmental conditions are more favorable, noting that the distribution of taxa is not static, depending on dynamic environmental conditions and therefore the current known distribution of a taxon does not always correspond to its historical distribution over time, since the current climate has not remained so for periods of time on scales relative to the processes of population genetics.
+(i.
+e.
+tens or hundreds of thousands of years).
+Even for some species these dynamics seem to induce ecological processes such as seasonal migrations in birds, or hibernation in mammals and frogs.
+But in general, environmental changes are reflected in geography as range expansions and contractions over time.
+That is, processes of population dynamics, driven by environmental fitness, in our approach.
+
+This change in population dynamics is implied by the genetic structure of populations through the concept of effective population size $N_e$ [@wright1931evolution; @karlin1968rates]. 
+So if we set a similar time scale for these changes, we will be able to observe the fitness relationship (through population dynamics) in both genetic structure and distribution range changes over time.
+That is, we need an index that reflects environmental changes in the geographic distribution of species and changes in the genetic structure of the population.
+In this case, we can consider the fixation index $F_{st}$ as an indicator of the difference between populations that come from a source population and that have undergone range shifts that we assume are driven by environmental changes.
+Under these assumptions we expect that it is possible to relate changes in the distribution of taxa to the genetic structure of their populations using statistical models to explain the relationship.
 

### 2023-07-11 manuscript_english.tex
expand introduction

@@
 \subsection{Introduction}\label{introduction}}
+
+Relating the genetic structure of the populations of a species to its
+geographic distribution is a complicated challenge from the point of
+view of biogeography. The task is made easier if evolutionary history is
+incorporated on both sides of the relationship. On the one hand, giving
+depth and an evolutionary perspective to the geographic distribution of
+a taxon relies primarily on ecological niche theory, and is driven by
+the massive amount of information currently available from both
+digitized biological collections and citizen science (through which
+records of taxa occurrence are through which records of taxa occurrence
+are available) as well as from environmental information available for
+geographic information systems and climate models. information systems
+and accurate climatological models of both the present and the past.
+
+This change in reproductive success is driven by a set of factors,
+however we propose in this work a conceptual framework where we consider
+only environmental factors at a macroecological scale, inspired by the
+concept of Grinnellian niches (those composed by sets of points of
+non-interactive variables in multidimensional spaces), seeking as a
+principle the necessary conditions for fitness and not the response of
+taxa. points of non-interactive variables in multidimensional spaces),
+seeking as a principle the conditions necessary for fitness and not the
+response of taxa.
+
+That is, we simplify our model (as is done in the physical sciences) to
+link through fitness the changes in biological populations and their
+structure, observed through population genetics, with patterns of
+spatial distribution over time. Thus, the goal is to link biogeographic
+patterns and evolutionary processes of populations in a single
+conceptual framework.
+
+It is from this Grinellian niche where we start, assuming that taxa in
+general have to be distributed in the geography where environmental
+conditions are more favorable, noting that the distribution of taxa is
+not static, depending on dynamic environmental conditions and therefore
+the current known distribution of a taxon does not always correspond to
+its historical distribution over time, since the current climate has not
+remained so for periods of time on scales relative to the processes of
+population genetics. (i. e. tens or hundreds of thousands of years).
+Even for some species these dynamics seem to induce ecological processes
+such as seasonal migrations in birds, or hibernation in mammals and
+frogs. But in general, environmental changes are reflected in geography
+as range expansions and contractions over time. That is, processes of
+population dynamics, driven by environmental fitness, in our approach.
+
+This change in population dynamics is implied by the genetic structure
+of populations through the concept of effective population size \(N_e\)
+(Wright 1931; Karlin 1968). So if we set a similar time scale for these
+changes, we will be able to observe the fitness relationship (through
+population dynamics) in both genetic structure and distribution range
+changes over time. That is, we need an index that reflects environmental
+changes in the geographic distribution of species and changes in the
+genetic structure of the population. In this case, we can consider the
+fixation index \(F_{st}\) as an indicator of the difference between
+populations that come from a source population and that have undergone
+range shifts that we assume are driven by environmental changes. Under
+these assumptions we expect that it is possible to relate changes in the
+distribution of taxa to the genetic structure of their populations using
+statistical models to explain the relationship.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wright1931evolution}{}}%
+Wright, Sewall. 1931. {``Evolution in Mendelian Populations.''}
+\emph{Genetics} 16 (2): 97.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%
-Wright, Sewall. 1943. {``Isolation by Distance.''} \emph{Genetics} 28
-(2): 114.
+---------. 1943. {``Isolation by Distance.''} \emph{Genetics} 28 (2):
+114.
 

### 2023-07-11 manuscript_english.tex
Changed small things, italics

@@
 That is, we simplify our model (as is done in the physical sciences) to
-link through fitness the changes in biological populations and their
+link through evolutionary fitness the changes in biological populations and their
 structure, observed through population genetics, with patterns of
@@
 
-For a given region \(\mathbf{G}\) in te geography, at any point in time,
+For a given region \(\mathbf{G}\) in the geography, at any point in time,
 we have a set of existing environmental conditions
@@
 depends on environmental conditions. If we take the set of places in
-environmental space where these values are greater than a threshold, we
+environmental space where these values are greater than a threshold (\textit{k}), we
 can define a volume in multidimensional space such that \(R(\vec{e})\)
@@
 expressing the fundamental niche as a function of environmental
-conditions. So we can correspong each point of geography to
+conditions. So we can corresponding each point of geography to
 environmental space, with different environmental conditions and
@@
 
-This equation was proposed by Soberon and Peterson Soberón and Peterson
+This equation was proposed by Soberón and Peterson
 (2020) to relate demographics to a function of fitness and environment.
@@
 existing niche) from known presence points in the geography
-\(\mathbf{G}\) obtained from emprical observations stored in database,
+\(\mathbf{G}\) obtained from empirical observations stored in database,
 and map to \(\mathbf{E}\) in order to numerically fitting a convex
@@
 and therefore the suitability function changes but not in a homogeneous
-way, so each site presents different displacement rates in envrionmental
+way, so each site presents different displacement rates in environmental
 spaces and as a consequence, different demographic rates. So we can
@@
 each subpopulation causes the average heterozygosity between populations
-(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)), i.e., for \(t=0\)
-\(F_{st} = 0\) and for very large \(t\) the index is \(1\). that is, for
-\(t=0\) \(F_{st} = 0\) and for very large \(t\) the index is \(1\). We
+(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)). That is, for
+\(t=0\) , \(F_{st} = 0\), and for very large \(t\) the index is \(1\). We
 have included a constant \(k\) without loss of generalization.
@@
 i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
-generation time), as we describe above, so we can asume an harmonic mean
+generation time), as we describe above, so we can assume an harmonic mean
 of suitabilities
 \(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
-substitue in the previous equation:
+substitute in the previous equation:
 
@@
 harmonic \emph{Suitability Prevalence Area}
-\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
+\(\mathbf{S} = \{S_i | S_i > s\}\), where \textit{s} is a threshold for the endemic historic area (see below), since we expect that areas where the
 \(F_{st}\) is lower are in areas where the prevalence over time is
@@
 \item
-  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
+  It has low speciation rate (0.25) and \textit{ca.} 5 million year divergence time from other squirrels (Upham, Esselstyn, and Jetz 2019); 
   therefore, we assume that its fundamental niche is stable over time.
@@
 
-\hypertarget{fixation-index-as-a-functino-of-spa}{%
-\subsubsection{Fixation index as a functino of
-SPA}\label{fixation-index-as-a-functino-of-spa}}
+\hypertarget{fixation-index-as-a-function-of-spa}{%
+\subsubsection{Fixation index as a function of
+SPA}\label{fixation-index-as-a-function-of-spa}}
 
@@
 structured populations, a matrix of fixation indices between populations
-is considered. For this work we considerer the data from @Bono et al.
+is considered. For this work we considerer the data from Bono et al.
 (2018). It consideres 10 populations along the known distribution and
@@
 \includegraphics{all_figures/figure_4.png}
-\caption{Suitability prevalence area. Here we observe the harmonic mean
-of suitability for each pixel along the distribucion from 0 to 1. The
+\caption{Suitability prevalence area of \textit{Sciurus aberti} over the last 120,000 years. Here we observe the harmonic mean
+of suitability for each pixel along the distribution from 0 to 1. The
 bold lines shows the the area when the harmonic mean es above 0.5
-threshold (blud line) and 0.9 threshold (black line)}
+threshold (blue line) and 0.9 threshold (black line)}
 \end{figure}
@@
 \caption{R squared from linear models (fixation index as a function of
-suitability prevalence area). The only R squared with statistical
-significance is barberi}\tabularnewline
+suitability prevalence area for the localities as listed in Bono et al. 2018). The only R squared with statistical
+significance is for \textit{S. a. barberi}.}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-S. a. aberti Carson-SFW & 0.000 & 0.962 \\
-S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
-S. a. aberti MT-Zuni & 0.002 & 0.894 \\
-S. a. aberti San Juan & 0.000 & 0.952 \\
-S. a. barberi & 0.794 & 0.001 \\
-S. a. chuscensis E & 0.002 & 0.892 \\
-S. a. chuscensis W & 0.018 & 0.713 \\
-S. a. ferreus Carson E & 0.040 & 0.581 \\
-S. a. ferreus Pike & 0.000 & 0.967 \\
-S. a. kaibabensis & 0.005 & 0.841 \\
+\textit{S. a. aberti} Carson-SFW & 0.000 & 0.962 \\
+\textit{S. a. aberti} Coconino-Gila & 0.131 & 0.304 \\
+\textit{S. a. aberti} MT-Zuni & 0.002 & 0.894 \\
+\textit{S. a. aberti} San Juan & 0.000 & 0.952 \\
+\textit{S. a. barberi} & 0.794 & 0.001 \\
+\textit{S. a. chuscensis} E & 0.002 & 0.892 \\
+\textit{S. a. chuscensis} W & 0.018 & 0.713 \\
+\textit{S. a. ferreus} Carson E & 0.040 & 0.581 \\
+\textit{S. a. ferreus} Pike & 0.000 & 0.967 \\
+\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
 \end{longtable}
@@
 \caption{Projection of the \(F_{st}\) model to the geography in the
-actual distribution of \emph{S aberti}}
+ current distribution of \emph{S. aberti}.}
 \end{figure}
+
+Discussion
 

### 2023-07-29 abstract.Rmd
changes in manuscript

--- 2023-07-11 manuscript_english.Rmd
@@
 bibliography: R21_citation.bib
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-The current distributions of species do not always correspond to their historical distributions over evolutionarily significant periods.
-This is because environmental conditions are not static over time; species tend to distribute where conditions are most favorable.
+The current distributions of species do not always correspond to their
+historical distributions over evolutionarily significant periods. This
+is because environmental conditions are not static over time; species
+tend to distribute where conditions are most favorable.
 
-It follows from this relationship between environment and species distribution that population size also varies with time.
-This variation in population size is related to the effective population size $N_e$ [@karlin1968rates], so that an index reflecting changes in environmental conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index $F_{st}$.
-Thus, it is possible to relate patterns of changes in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.
+It follows from this relationship between environment and species
+distribution that population size also varies with time. This variation in population size is related to the effective population size $N_e$, so that an index reflecting changes in environmental
+conditions in geography might be expected to be related to $N_e$ and
+thus to indicators of population structure such as the fixation index
+$F_{st}$. Thus, it is possible to relate patterns of changes in the
+distribution of species to the genetic structure of their populations
+using a statistical model that explains this relationship.
 
-With this approach we can predict the geographic pattern of population structure from environmental information.
-This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting] and population genetics theories [@lira2014relationship].
+With this approach we can predict the geographic pattern of population
+structure from environmental information. This approach is strongly
+driven by advances in currently available climate simulations
+[@leonardi2023pastclim; @krapp2021statistics], as well as next
+generation sequencing data, and supported by both ecological niche
+[@thorup2021response; @nogues2009predicting] and population genetics
+theories [@lira2014relationship].
 
@@
 
-To obtain the SPA, we performed a historical reconstruction of the geographical range back to 120 000 BC at 2 000 year intervals and recorded the environmental suitability at each site.
+To obtain the SPA, we performed a historical reconstruction of the
+geographical range back to 120 000 BC at 2 000 year intervals and
+recorded the environmental suitability at each site.
 
-Subsequently, we delimited historical endemic areas to locations where the prevalence of suitability remained at 90% during this period.
-From the fixation index calculated with respect to populations in the historical endemic areas, a statistical model of the fixation index as a function of SPA was performed.
+Subsequently, we delimited historical endemic areas to locations where
+the prevalence of suitability remained at 90\% during this period. From the fixation index calculated with respect to populations in the
+historical endemic areas, a statistical model of the fixation index as a function of SPA was performed.
 With this statistical model, the fixation index values were projected to the current distribution to obtain a map with the geographic pattern of this index.
 
-As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
+As a case study we consider squirrels (*Sciurus aberti*), a species
+currently distributed in disjunct patches from the southern Rocky
+Mountains in the United States to the northern Sierra Madre Occidental
+in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
 
-Our results reveal that suitability prevalence corresponds to the fixation index of *S.
-aberti* populations with respect to a source population.
+Our results reveal that suitability prevalence corresponds to the
+fixation index of *S. aberti* populations with respect to a source
+population.
 Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value.
 Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
-
-## Introduction
-
-### Theoretical basis
-
-#### Environment, fitness and suitability
-
-For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$.
-We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$.
-Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions.
-If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$.
-i.
-e.
-the set of environmental conditions for which values of fitness are capable of sustaining populations.
-We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$.
-This is a way of expressing the fundamental niche as a function of environmental conditions.
-So we can correspong each point of geography to environmental space,  with different environmental conditions and different growth rates.
-However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$.
-Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.
-
-This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of fitness and environment.
-In that study they tested the validity with monthly variables and empirical values on a short time scale.
-relating life history parameters and the environmental suitability of niche models.
-
-This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$.
-In a practical way, instead of $\mathbf{N}_F$, we approximated
-$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
-(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.
-
-The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $S_i(\vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
-
-Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
-$$
-N_i \propto  S_i(\vec{e}_i)
-$$
-
-![Hypothetical niche in environmental space.
-An arbitrary point moves within the environmental space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
-
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates.
-So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:
-
-$$
-N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
-$$
-
-#### Genetic variation and effective population size
-
-On the other hand, loss of heterozygosity may be related to the amount of genetic variation present in the absence of mutation and selection, so theoretically, one would expect a strong correlation between effective population size and heterozygosity on the basis of population genetics.
-We start from the assumption that in the absence of gene flow between subpopulations, the rate of fixation by isolation depends on the effective population size $N_e$ when subpopulations diverge from a common ancestor in $t$ generations.
-[@hedrick2005standardized: @wright1943isolation]:
-
-$$
-F_{st}(t) = 1 - e^{-kt/N_e } 
-$$
-
-where we assume, according to Nei [@nei1986definition], that genetic drift within each subpopulation causes the average heterozygosity between populations ($H_s$) to approach zero ($F_{st} = 1 - H_s/H_t$), i.e., for $t=0$ $F_{st} = 0$ and for very large $t$ the index is $1$.
-that is, for $t=0$ $F_{st} = 0$ and for very large $t$ the index is $1$. We have included a constant $k$ without loss of generalization.
-
-Using $e(x) = \sum_{k=0}^{\infty} \frac{x^k}{k!} = 1 + x + \frac{x^2}{2} + ...$ and leaving to first order for $N_e$ large enough
-
-$$
-F_{st} \approx \frac{kt}{N_e} + C
-$$
-
-Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ...
-+ \frac{1}{N_\tau} } \Big)$ for each $i$ site in $\mathbf{G}$.
-We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close:
-
-$$
-F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
-$$
-
-#### Relationship between the fixation index and environmental suitability
-We can infer locally for each population $N_i$ in the geography expect to be proportional to the environmental suitability for each time considered, 
-For simplification in the index we select a $N_i$ population in the geography and infer that locally, this population is proportional to the environmental suitability in each time considered, i.
-e.
-$N_{j} \propto S_{j}(\vec{e}_{j})$  (each index $j$ is for the generation time), as we describe above, so we can asume an harmonic mean of suitabilities $\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j}  = \frac{1}{S}$ and substitue in the previous equation:
-
-$$
-F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{S_j(\vec{e_j})}
-$$
-
-We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time.
-Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the suitability prevalence for an $i$ site in the geography:
-
-$$
-F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
-$$
-If we consider all sites in the available geography, we call this harmonic  *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
-environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred.
-Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.
-
-## Methods
-
-We summarize our 5 steps predictive approach in the figure 2.
-
-<p align="center">
-![Realized niche in environmental space. A point dynamically moves along different time slices](all_figures/figure_2.png){width=30%}
-</p>
-
-We propose a method to find the Suitability Prevalence Area (SPA) as an index with a dual purpose: 1) to find endemic areas to delimit the historical distribution of species and 2) to explain patterns of genetic diversity.
-
-The method for estimating SPA consists of a series of 4 steps:
-
-1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
-2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
-3. Characterize past environments and retain projections only for sites with analogous environmental conditions to the present.
-4. Calculate the harmonic mean of the suitability for all the sites all time slices.
-
-Subsequently, one application of the SPA is to explain genetic diversity (the fixation index) as a function of the SPA and project it to the geography.
-For this it is necessary to perform a series of steps, which consist of:
-
-- Obtain georeferenced points of genetically structured populations with fixation index values.
-- Extract values from the SPA map with these points. 
-- A linear model of the fixation index value with respect to the SPA is performed.
-- With the linear model all SPA values of suitability are interpolated.
-- A map is generated with all the values interpolated by the model.
-
-To test our model we use the Aberti squirrel (*Sciurus aberti*) as a case study because:
-
-- Since it is a medium-sized mammal (0.5 kg) its identification is easy.
-- The subspecies are geographically isolated, since they are found in pine forests surrounded by desert. Therefore we can consider the assumption of isolation by distance [@wright1943isolation]. 
-- There are sufficient occurrence records for the entire range of the known distribution.
-- It has low speciation rate (0.25) [@upham2019inferring], therefore, we assume that its fundamental niche is stable over time.
-- There is fixation index information [@bono2018genome] for different populations of the subspecies, which is useful for using SPA as a way to explain genetic differentiation between populations.
-
-### Unique occurrence records 
-We have downloaded geographic information on Sciurus aberti from the open access platform Global Biodiversity Information Facility (GBIF).
-Subsequently, we eliminated duplicate records, those without precise coordinates and those with uncorroborated collection information.
-Once the unique occurrence records were obtained, we filtered the geographic space in order to avoid overpredictions in the spatial distribution models caused by the agglomeration of occurrence points.
-The filtering consisted of superimposing the records in the geographic space, then we made 5 latitudinal slices of 5 degrees of separation between each one, from latitude 20 degrees south to latitude 45 degrees north, taking as criteria the known distribution of the species and the ecoregions of North America, level 1 (downloaded from the US Environmental Protection Agency, 2010).
-Once the geographic space was segmented, we left the same number of records for each band (30 records per band) obtaining a total of 150 unique records of presence.
-
-### Environmental characterization
-The environmental information to generate the models, from the present and the reconstruction to the past, was obtained from Pastclim 1.2, an R statistical software package designed to download and manipulate paleoclimatic datasets.
-In our particular case we chose the Beyer at al.
-@beyer2020high set as it has environmental information available up to 120 000 BE, significantly evolutionary time, at 2000 year time intervals based on Global Circulation Models HadCM3 [@singarayer2010high] and HadAM3H [@valdes2017bridge].
-The environmental dataset has 17 bias-corrected bioclimatic variables, reduced spatial scale and spatial resolution of 0.5°square cells.
-
-### Accessibility region ($\textbf{M}$).
-Species distribution models according to the BAM diagram [@soberon2005interpretation] states that the geographic range occupied by a species ($\textbf{G}_o$) is the region of appropriate assemblages in terms of abiotic conditions.
-The region $\textbf{M}$ represents in geography areas in which the species has access due to its colonization capabilities and the structure of geographic barriers within a specific period of time [@soberon2009niches].
-
-In this study we selected a set of North American ecoregions for modeling to delimit our Accessibility region.
-We took into account ecoregions that overlap with the known distribution of the species and also took into account neighboring ecoregions to increase the potential accessibility in projection scenarios to the past.
-Thus, regions 5, 6, 9, 10, 12, 13 of the level I ecoregions (<https://www.epa.gov/>) were selected.
-
-### Niche models 
-#### Current distribution
-Estimating niche boundaries for species occurrence is called ecological niche modeling (ENM), and when the emphasis is on geographic distribution, it is known as species distribution modeling (SDM) [@guisan2005predicting; @peterson2012species; @saupe2012variation].
-We model the current distribution of S.
-aberti using the MAXENT 3.4.4 algorithm [@phillips2006maximum], which is a correlative model based on the maximum entropy principle used to estimate species distributions.
-The output of Maxent is a relative occurrence rate (ROR) interpreted as a probability of habitat suitability given observed environmental conditions by presence-only points [@merow2013practical].
-
-The parameterization was performed with the default values of the program with the exception of the "extrapolate" and "clamping" options to avoid artificial extrapolations in the extreme values of the climatic variables used in the models.
-To calibrate the models we used 70\% of the records and the remaining 30\% to validate them.
-
-In order to give statistical certainty and to consider uncertainty, as well as to reduce overfitting and to have a suitable model to extrapolate to past scenarios, 10 replications were performed for each model using a cross-validation with the training data.
-In the end, the average model of these replications was considered as the result.
-
-We evaluated the statistical significance of the resulting models for the present using the partial ROC test [@peterson2008rethinking], which is a modification of the ROC (receiver Operation Characteristic) test.
-The results of this test are proportions (ratios) of the area under the curve of the model with respect to a null model, product of repetitions that allow to statistically evaluate the areas under the curve (AUC) in relation to that expected by chance [@peterson2008rethinking], where a value derived by chance would be $0$ and an acceptable value, according to the proportion of minimum omission errors tolerated in the model, would be greater than $1$.
-A different set of test data was used on the average model of the replicates and in this way the reduction of overfitting was guaranteed.
-
-#### Reconstruction to the past
-The average model obtained from Maxent was extrapolated to environmental conditions into the past.
-Environmental layers were taken from pastclim [@leonardi2023pastclim] in 2000 year time slices to cover a total of 120 000 years.
-For each scenario into the past, 10 replicates were cross-validated and averaged.
-
-#### Analogy of environmental conditions 
-To be certain that areas in the past are climatically analogous to the present we implemented the mobility-orient parity (MOP) test.
-This method generates a value from $0$ to $1$ according to the environmental analogy of past conditions to the present to eliminate geographic areas in the projections to the past outside of climates analogous to the present [@owens2013constraints].
-
-### Suitability Prevalence Area (SPA)
-Finally, to generate the SPA, both the average projection of the present and all the average projections of the past are considered. The output files are read in raster format.
-
-A harmonic mean is performed as follows:
-For each layer the inverse of each cell value is obtained.
-Then the inverse values are summed by cells and divided by the number of scenarios.
-Finally, the inverse of each cell is obtained.
-It is important to consider that the Maxent outputs may contain very small values close to zero in many cells, for this reason we added a small amount of $0.001$ to all the values in all the maps so that when obtaining the harmonic mean we could have an inverse value for values close to zero without altering the biological sense of environmental suitability.
-
-### Endemic historic Area
-To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $\mathbf{S}$ equation.
-This threshold is the proportion of time that pixel has had ideal conditions.
-For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.
-
-### Fixation index as a functino of SPA
-
-To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered.
-For this work we considerer the data from @@bono2018genome.
-It consideres 10 populations along the known distribution and provides values of $F_{st}$ between this populations in matrix form.
-The matrix was transformed into a table as pairs between populations with index values between each pair.
-The values of the population with itself, i.e.
-$0$, were included.
-Subsequently, georeferences were obtained for each population considered in the fixation index matrix.
-From this geo-referencing the SPA value for each population was extracted.
-Finally, the SPA value was aligned in the table of fixation index pairs.
-In total the table has 4 columns: source population, target population, fixation index and SPA of the target population.
-This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$.
-An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.
-
-### Fixation index projected in geography
-The linear model that presents the highest value of $R^2$ is considered suitable to make a geographic projection of the fixation index and observe genetic diversity in geographic space.
-To generate this map, the SPA values are taken and a table of coordinates and SPA values is generated for each cell of the raster.
-The SPA values are then extrapolated to obtain fixation index values from the linear model obtained previously.
-This column is added to the coordinate table and a raster is generated with the projected $F_{st}$ values.
-Finally, a cutoff of this raster is performed using a map of the distribution of the species at present with a cutoff threshold of 90% in binary format (where $1 =$presence and $0 =$ absence), since the records of the presence localities were obtained from bases that might contain some errors in the data [@phillips2006maximum].
-
-## Results
-
-### Current and past distribution
-We show the species distribution model for *Sciurus aberti*  along the time line of the projected area per time slice. We show 5 notable moments in time line. 
-and 5 notable points 
-<p align="center">
-![A) Species distribution model for *Sciurus aberti* present distribution.
-B) Relative projected area for SDM in pixels acoording a threshold of 0.9 suitability for each past time-sliced environmental conditions.
-We can observe a maximum area corresponding to the last interglacial era and an inverted U-shape in the trend of the area along the time considered.
-Also It can be noted that the actual distribution has similar area to the coditions 120 000 years ago.
-B) and D) corresponds to distribution peaks along the time scale and C) and E) corresponds to minimum  ](all_figures/figure_3.png){width=100%}
-</p>
-
-### SPA and Endemic Historic Area
-We get the suitability presence area (figure 2) delmited by a 0.5 and 0.9 threshold (Figure 4).
-The endemic region where the enviornmental conditions 90 \% del tiempo is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango.
-We correspond this Area to the *Sciurus aberti barbieri* subspecies.
-Also
-is notable that at 0.5 threshold, other regions seems to have stable conditions.
-We observe the area close to Coronado National Forest as well as Gila National Forest and Mescalero Reservation as suitable areas along time in the United States and some Areas of the Volcanic Belt in Mexico (Puebla, north of Guerrero State and Jalisco.
-<p align="center">
-![Suitability prevalence area.
-Here we observe the harmonic mean of suitability for each pixel along the distribucion from 0 to 1.
-The bold lines shows the the area when the harmonic mean es above 0.5 threshold (blud line) and 0.9 threshold (black line)](all_figures/figure_4.png){}
-</p>
-
-### Fst as a funciton of SPA and Fixation index projected in geography
-
-For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an R-squared value of 0.794.
-
-```{r, echo=FALSE}
-aberti_pops <- read.csv(file = "all_tables/fst_SPA_rsquared.csv")
-
-knitr::kable(aberti_pops, "simple", align = "ccc", caption = "R squared from linear models (fixation index as a function of suitability prevalence area).
-The only R squared with statistical significance is barberi")
-```
-
-It can be noted that S.
-a.
-aberti Coconino-Gila had an R-squared value greater than 0 (0.131) but it is not statistically significant.
-Furthermore, in our approach we imputed the value of the origin with an Fst of zero.
-To show that the model hypothesis still holds, we performed a regression for barberi without the point of origin imposed (Figure 5b), in addition to eliminating the S.
-aberti chuscensis populations due to their low uncertainty in georeference, obtaining an R-squared of 0.72 (p value = 0.016).
-
-```{r, figures-side, fig.show="hold", out.width="50%",  fig.cap="Linear regression for S a.
-barberi population A) SPA in all populations B) withouth barberi and chusensis populations.
-We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)", echo=FALSE}
-knitr::include_graphics(c("all_figures/figure_5a.png","all_figures/figure_5b.png"))
-```
-
-From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S. a. barberi* population. (Figure 6)
-
-<p align="center">
-![Projection of the $F_{st}$ model to the geography in the actual distribution of *S aberti*](all_figures/figure_6.png){width=80%}
-</p>
-
-### References

### 2023-07-29 abstract.tex
changes in manuscript

--- 2023-07-11 manuscript_english.tex
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 distribution that population size also varies with time. This variation
-in population size is related to the effective population size \(N_e\)
-(Karlin 1968), so that an index reflecting changes in environmental
-conditions in geography might be expected to be related to \(N_e\) and
-thus to indicators of population structure such as the fixation index
+in population size is related to the effective population size \(N_e\),
+so that an index reflecting changes in environmental conditions in
+geography might be expected to be related to \(N_e\) and thus to
+indicators of population structure such as the fixation index
 \(F_{st}\). Thus, it is possible to relate patterns of changes in the
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
-\hypertarget{theoretical-basis}{%
-\subsubsection{Theoretical basis}\label{theoretical-basis}}
-
-\hypertarget{environment-fitness-and-suitability}{%
-\paragraph{Environment, fitness and
-suitability}\label{environment-fitness-and-suitability}}
-
-For a given region \(\mathbf{G}\) in te geography, at any point in time,
-we have a set of existing environmental conditions
-\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
-We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
-environmental variables, that is, a point in the space of \(v\)
-environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
-\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
-that is, for a species with non-overlapping generations, the
-population's net reproductive rate \(R_0\) is a measure of fitness that
-depends on environmental conditions. If we take the set of places in
-environmental space where these values are greater than a threshold, we
-can define a volume in multidimensional space such that \(R(\vec{e})\)
-is always positive and greater than this threshold, that is
-\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
-environmental conditions for which values of fitness are capable of
-sustaining populations. We call
-\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
-expressing the fundamental niche as a function of environmental
-conditions. So we can correspong each point of geography to
-environmental space, with different environmental conditions and
-different growth rates. However all sites within the region defined by
-\(\mathbf{N}_F\) will have viable populations and positive rates
-depending on time because we assume that the environment is dynamic and
-each geographic point is static, so for each site in geographic space
-this growth rate depends on the environment and therefore is different
-at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
-will be related to the local population size as long as it is within the
-boundary defined by \(\mathbf{N}_F\).
-
-This equation was proposed by Soberon and Peterson Soberón and Peterson
-(2020) to relate demographics to a function of fitness and environment.
-In that study they tested the validity with monthly variables and
-empirical values on a short time scale. relating life history parameters
-and the environmental suitability of niche models.
-
-This set can be seen as a convex volume in environmental space and we
-relate this growth rate to an environmental suitability function bounded
-between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
-it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any
-function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
-\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
-values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
-practical way, instead of \(\mathbf{N}_F\), we approximated
-\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
-existing niche) from known presence points in the geography
-\(\mathbf{G}\) obtained from emprical observations stored in database,
-and map to \(\mathbf{E}\) in order to numerically fitting a convex
-volume of the environmental space that enclose the known information.
-
-The objective of our approach is not to know the function of \(R_0\) but
-rather to relate this local variation and the function \(\mathbf{N}_F\)
-with a function \(S_i(\vec{e}_i)\) of environmental suitability that can
-be estimated from known methods of applied mathematics and machine
-learning (for example the case of Maxent, Phillips, Anderson, and
-Schapire (2006)).
-
-Population size locally is a function of environmental suitability,
-where if suitability is \(0\) the population is \(0\) and if suitability
-is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) \[
-N_i \propto  S_i(\vec{e}_i)
-\]
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
-\caption{Hypothetical niche in environmental space. An arbitrary point
-moves within the environmental space entering and exiting the volume
-(niche) where the intrinsic growth rate is positive.}
-\end{figure}
-
-Once we have an approximation of this volume, we consider that each
-point in the geography is static and has a trajectory in the
-environmental space and for each trajectory the fitness function changes
-and therefore the suitability function changes but not in a homogeneous
-way, so each site presents different displacement rates in envrionmental
-spaces and as a consequence, different demographic rates. So we can
-establish that for each site \(i\) in geographic space \(\mathbf{G}\),
-there is a population size \(N_i\) that changes with respect to a
-fitness function dependent on time-varying environmental conditions:
-
-\[
-N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
-\]
-
-\hypertarget{genetic-variation-and-effective-population-size}{%
-\paragraph{Genetic variation and effective population
-size}\label{genetic-variation-and-effective-population-size}}
-
-On the other hand, loss of heterozygosity may be related to the amount
-of genetic variation present in the absence of mutation and selection,
-so theoretically, one would expect a strong correlation between
-effective population size and heterozygosity on the basis of population
-genetics. We start from the assumption that in the absence of gene flow
-between subpopulations, the rate of fixation by isolation depends on the
-effective population size \(N_e\) when subpopulations diverge from a
-common ancestor in \(t\) generations. Wright (1943):
-
-\[
-F_{st}(t) = 1 - e^{-kt/N_e } 
-\]
-
-where we assume, according to Nei (Nei 1986), that genetic drift within
-each subpopulation causes the average heterozygosity between populations
-(\(H_s\)) to approach zero (\(F_{st} = 1 - H_s/H_t\)), i.e., for \(t=0\)
-\(F_{st} = 0\) and for very large \(t\) the index is \(1\). that is, for
-\(t=0\) \(F_{st} = 0\) and for very large \(t\) the index is \(1\). We
-have included a constant \(k\) without loss of generalization.
-
-Using
-\(e(x) = \sum_{k=0}^{\infty} \frac{x^k}{k!} = 1 + x + \frac{x^2}{2} + ...\)
-and leaving to first order for \(N_e\) large enough
-
-\[
-F_{st} \approx \frac{kt}{N_e} + C
-\]
-
-Moreover, the effective population can be estimated as the harmonic mean
-of the population in a demographic time series (Karlin 1968) for
-\(\tau\) generations
-\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\)
-for each \(i\) site in \(\mathbf{G}\). We substitute the harmonic mean
-into the above equation, considering that although the number of
-generations is not exactly equal to the time of the fixation index, they
-are expected to be very close:
-
-\[
-F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
-\]
-
-\hypertarget{relationship-between-the-fixation-index-and-environmental-suitability}{%
-\paragraph{Relationship between the fixation index and environmental
-suitability}\label{relationship-between-the-fixation-index-and-environmental-suitability}}
-
-We can infer locally for each population \(N_i\) in the geography expect
-to be proportional to the environmental suitability for each time
-considered, For simplification in the index we select a \(N_i\)
-population in the geography and infer that locally, this population is
-proportional to the environmental suitability in each time considered,
-i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
-generation time), as we describe above, so we can asume an harmonic mean
-of suitabilities
-\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
-substitue in the previous equation:
-
-\[
-F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{S_j(\vec{e_j})}
-\]
-
-We call this harmonic mean of environmental suitability
-\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
-\(1\), since it reflects the rate of population growth as a function of
-environment over a period of time. Thus we have a way to check that the
-loss of heterozygosity (through the fixation index \(F_{st}\)) is
-inversely proportional to the suitability prevalence for an \(i\) site
-in the geography:
-
-\[
-F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
-\] If we consider all sites in the available geography, we call this
-harmonic \emph{Suitability Prevalence Area}
-\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
-\(F_{st}\) is lower are in areas where the prevalence over time is
-higher, i.e.~environmental conditions remain suitable for a significant
-period of time and therefore no effects of genetic drift of populations
-have occurred. Conversely, it is expected that in populations with a
-high fixation index value, fitness will not have prevailed over the same
-time interval even though conditions currently exist and populations are
-present.
-
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-We summarize our 5 steps predictive approach in the figure 2.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.3\textwidth,height=\textheight]{all_figures/figure_2.png}
-\caption{Realized niche in environmental space. A point dynamically
-moves along different time slices}
-\end{figure}
-
-We propose a method to find the Suitability Prevalence Area (SPA) as an
-index with a dual purpose: 1) to find endemic areas to delimit the
-historical distribution of species and 2) to explain patterns of genetic
-diversity.
-
-The method for estimating SPA consists of a series of 4 steps:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Model the potential distribution of a taxon with present-day
-  environmental conditions and obtain an environmental suitability map.
-\item
-  Project the potential distribution model to various environmental
-  scenarios in the past with constant time steps and obtain the
-  environmental suitability for each time scenario.
-\item
-  Characterize past environments and retain projections only for sites
-  with analogous environmental conditions to the present.
-\item
-  Calculate the harmonic mean of the suitability for all the sites all
-  time slices.
-\end{enumerate}
-
-Subsequently, one application of the SPA is to explain genetic diversity
-(the fixation index) as a function of the SPA and project it to the
-geography. For this it is necessary to perform a series of steps, which
-consist of:
-
-\begin{itemize}
-\tightlist
-\item
-  Obtain georeferenced points of genetically structured populations with
-  fixation index values.
-\item
-  Extract values from the SPA map with these points.
-\item
-  A linear model of the fixation index value with respect to the SPA is
-  performed.
-\item
-  With the linear model all SPA values of suitability are interpolated.
-\item
-  A map is generated with all the values interpolated by the model.
-\end{itemize}
-
-To test our model we use the Aberti squirrel (\emph{Sciurus aberti}) as
-a case study because:
-
-\begin{itemize}
-\tightlist
-\item
-  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
-\item
-  The subspecies are geographically isolated, since they are found in
-  pine forests surrounded by desert. Therefore we can consider the
-  assumption of isolation by distance (Wright 1943).
-\item
-  There are sufficient occurrence records for the entire range of the
-  known distribution.
-\item
-  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
-  therefore, we assume that its fundamental niche is stable over time.
-\item
-  There is fixation index information (Bono et al. 2018) for different
-  populations of the subspecies, which is useful for using SPA as a way
-  to explain genetic differentiation between populations.
-\end{itemize}
-
-\hypertarget{unique-occurrence-records}{%
-\subsubsection{Unique occurrence
-records}\label{unique-occurrence-records}}
-
-We have downloaded geographic information on Sciurus aberti from the
-open access platform Global Biodiversity Information Facility (GBIF).
-Subsequently, we eliminated duplicate records, those without precise
-coordinates and those with uncorroborated collection information. Once
-the unique occurrence records were obtained, we filtered the geographic
-space in order to avoid overpredictions in the spatial distribution
-models caused by the agglomeration of occurrence points. The filtering
-consisted of superimposing the records in the geographic space, then we
-made 5 latitudinal slices of 5 degrees of separation between each one,
-from latitude 20 degrees south to latitude 45 degrees north, taking as
-criteria the known distribution of the species and the ecoregions of
-North America, level 1 (downloaded from the US Environmental Protection
-Agency, 2010). Once the geographic space was segmented, we left the same
-number of records for each band (30 records per band) obtaining a total
-of 150 unique records of presence.
-
-\hypertarget{environmental-characterization}{%
-\subsubsection{Environmental
-characterization}\label{environmental-characterization}}
-
-The environmental information to generate the models, from the present
-and the reconstruction to the past, was obtained from Pastclim 1.2, an R
-statistical software package designed to download and manipulate
-paleoclimatic datasets. In our particular case we chose the Beyer at al.
-Beyer, Krapp, and Manica (2020) set as it has environmental information
-available up to 120 000 BE, significantly evolutionary time, at 2000
-year time intervals based on Global Circulation Models HadCM3
-(Singarayer and Valdes 2010) and HadAM3H (Valdes et al. 2017). The
-environmental dataset has 17 bias-corrected bioclimatic variables,
-reduced spatial scale and spatial resolution of 0.5°square cells.
-
-\hypertarget{accessibility-region-textbfm.}{%
-\subsubsection{\texorpdfstring{Accessibility region
-(\(\textbf{M}\)).}{Accessibility region (\textbackslash textbf\{M\}).}}\label{accessibility-region-textbfm.}}
-
-Species distribution models according to the BAM diagram (Soberon and
-Peterson 2005) states that the geographic range occupied by a species
-(\(\textbf{G}_o\)) is the region of appropriate assemblages in terms of
-abiotic conditions. The region \(\textbf{M}\) represents in geography
-areas in which the species has access due to its colonization
-capabilities and the structure of geographic barriers within a specific
-period of time (Soberón and Nakamura 2009).
-
-In this study we selected a set of North American ecoregions for
-modeling to delimit our Accessibility region. We took into account
-ecoregions that overlap with the known distribution of the species and
-also took into account neighboring ecoregions to increase the potential
-accessibility in projection scenarios to the past. Thus, regions 5, 6,
-9, 10, 12, 13 of the level I ecoregions (\url{https://www.epa.gov/})
-were selected.
-
-\hypertarget{niche-models}{%
-\subsubsection{Niche models}\label{niche-models}}
-
-\hypertarget{current-distribution}{%
-\paragraph{Current distribution}\label{current-distribution}}
-
-Estimating niche boundaries for species occurrence is called ecological
-niche modeling (ENM), and when the emphasis is on geographic
-distribution, it is known as species distribution modeling (SDM) (Guisan
-and Thuiller 2005; Peterson and Soberón 2012; Saupe et al. 2012). We
-model the current distribution of S. aberti using the MAXENT 3.4.4
-algorithm (Phillips, Anderson, and Schapire 2006), which is a
-correlative model based on the maximum entropy principle used to
-estimate species distributions. The output of Maxent is a relative
-occurrence rate (ROR) interpreted as a probability of habitat
-suitability given observed environmental conditions by presence-only
-points (Merow, Smith, and Silander Jr 2013).
-
-The parameterization was performed with the default values of the
-program with the exception of the ``extrapolate'' and ``clamping''
-options to avoid artificial extrapolations in the extreme values of the
-climatic variables used in the models. To calibrate the models we used
-70\% of the records and the remaining 30\% to validate them.
-
-In order to give statistical certainty and to consider uncertainty, as
-well as to reduce overfitting and to have a suitable model to
-extrapolate to past scenarios, 10 replications were performed for each
-model using a cross-validation with the training data. In the end, the
-average model of these replications was considered as the result.
-
-We evaluated the statistical significance of the resulting models for
-the present using the partial ROC test (Peterson, Papeş, and Soberón
-2008), which is a modification of the ROC (receiver Operation
-Characteristic) test. The results of this test are proportions (ratios)
-of the area under the curve of the model with respect to a null model,
-product of repetitions that allow to statistically evaluate the areas
-under the curve (AUC) in relation to that expected by chance (Peterson,
-Papeş, and Soberón 2008), where a value derived by chance would be \(0\)
-and an acceptable value, according to the proportion of minimum omission
-errors tolerated in the model, would be greater than \(1\). A different
-set of test data was used on the average model of the replicates and in
-this way the reduction of overfitting was guaranteed.
-
-\hypertarget{reconstruction-to-the-past}{%
-\paragraph{Reconstruction to the
-past}\label{reconstruction-to-the-past}}
-
-The average model obtained from Maxent was extrapolated to environmental
-conditions into the past. Environmental layers were taken from pastclim
-(Leonardi et al. 2023) in 2000 year time slices to cover a total of 120
-000 years. For each scenario into the past, 10 replicates were
-cross-validated and averaged.
-
-\hypertarget{analogy-of-environmental-conditions}{%
-\paragraph{Analogy of environmental
-conditions}\label{analogy-of-environmental-conditions}}
-
-To be certain that areas in the past are climatically analogous to the
-present we implemented the mobility-orient parity (MOP) test. This
-method generates a value from \(0\) to \(1\) according to the
-environmental analogy of past conditions to the present to eliminate
-geographic areas in the projections to the past outside of climates
-analogous to the present (Owens et al. 2013).
-
-\hypertarget{suitability-prevalence-area-spa}{%
-\subsubsection{Suitability Prevalence Area
-(SPA)}\label{suitability-prevalence-area-spa}}
-
-Finally, to generate the SPA, both the average projection of the present
-and all the average projections of the past are considered. The output
-files are read in raster format.
-
-A harmonic mean is performed as follows: For each layer the inverse of
-each cell value is obtained. Then the inverse values are summed by cells
-and divided by the number of scenarios. Finally, the inverse of each
-cell is obtained. It is important to consider that the Maxent outputs
-may contain very small values close to zero in many cells, for this
-reason we added a small amount of \(0.001\) to all the values in all the
-maps so that when obtaining the harmonic mean we could have an inverse
-value for values close to zero without altering the biological sense of
-environmental suitability.
-
-\hypertarget{endemic-historic-area}{%
-\subsubsection{Endemic historic Area}\label{endemic-historic-area}}
-
-To calculate the historic endemic area, the SPA result is taken and a
-threshold is applied to generate a binary map according to the
-\(\mathbf{S}\) equation. This threshold is the proportion of time that
-pixel has had ideal conditions. For our study, we consider the value of
-90\% since for a period of 120 000 years it is a coverage of 100 000
-years, which is a significantly evolutionary time to fix genetic
-patterns at the population level.
-
-\hypertarget{fixation-index-as-a-functino-of-spa}{%
-\subsubsection{Fixation index as a functino of
-SPA}\label{fixation-index-as-a-functino-of-spa}}
-
-To test our hypothesis of genetic diversity as a function of
-environmental suitability prevalence, from a set of genetically
-structured populations, a matrix of fixation indices between populations
-is considered. For this work we considerer the data from @Bono et al.
-(2018). It consideres 10 populations along the known distribution and
-provides values of \(F_{st}\) between this populations in matrix form.
-The matrix was transformed into a table as pairs between populations
-with index values between each pair. The values of the population with
-itself, i.e.~\(0\), were included. Subsequently, georeferences were
-obtained for each population considered in the fixation index matrix.
-From this geo-referencing the SPA value for each population was
-extracted. Finally, the SPA value was aligned in the table of fixation
-index pairs. In total the table has 4 columns: source population, target
-population, fixation index and SPA of the target population. This table
-was separated by source population groups to generate a set of linear
-models between SPA values and fixation indexes \(F_{st}\). An analysis
-was performed for each source population with the hypothesis that the
-correlation between SPA and fixation index exists and is non-zero if a
-dispersal process has occurred from the historical endemic area of the
-taxon.
-
-\hypertarget{fixation-index-projected-in-geography}{%
-\subsubsection{Fixation index projected in
-geography}\label{fixation-index-projected-in-geography}}
-
-The linear model that presents the highest value of \(R^2\) is
-considered suitable to make a geographic projection of the fixation
-index and observe genetic diversity in geographic space. To generate
-this map, the SPA values are taken and a table of coordinates and SPA
-values is generated for each cell of the raster. The SPA values are then
-extrapolated to obtain fixation index values from the linear model
-obtained previously. This column is added to the coordinate table and a
-raster is generated with the projected \(F_{st}\) values. Finally, a
-cutoff of this raster is performed using a map of the distribution of
-the species at present with a cutoff threshold of 90\% in binary format
-(where \(1 =\)presence and \(0 =\) absence), since the records of the
-presence localities were obtained from bases that might contain some
-errors in the data (Phillips, Anderson, and Schapire 2006).
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\hypertarget{current-and-past-distribution}{%
-\subsubsection{Current and past
-distribution}\label{current-and-past-distribution}}
-
-We show the species distribution model for \emph{Sciurus aberti} along
-the time line of the projected area per time slice. We show 5 notable
-moments in time line. and 5 notable points
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,height=\textheight]{all_figures/figure_3.png}
-\caption{A) Species distribution model for \emph{Sciurus aberti} present
-distribution. B) Relative projected area for SDM in pixels acoording a
-threshold of 0.9 suitability for each past time-sliced environmental
-conditions. We can observe a maximum area corresponding to the last
-interglacial era and an inverted U-shape in the trend of the area along
-the time considered. Also It can be noted that the actual distribution
-has similar area to the coditions 120 000 years ago. B) and D)
-corresponds to distribution peaks along the time scale and C) and E)
-corresponds to minimum}
-\end{figure}
-
-\hypertarget{spa-and-endemic-historic-area}{%
-\subsubsection{SPA and Endemic Historic
-Area}\label{spa-and-endemic-historic-area}}
-
-We get the suitability presence area (figure 2) delmited by a 0.5 and
-0.9 threshold (Figure 4). The endemic region where the enviornmental
-conditions 90 \% del tiempo is the Sierra Madre Occidental, in the
-limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this
-Area to the \emph{Sciurus aberti barbieri} subspecies. Also is notable
-that at 0.5 threshold, other regions seems to have stable conditions. We
-observe the area close to Coronado National Forest as well as Gila
-National Forest and Mescalero Reservation as suitable areas along time
-in the United States and some Areas of the Volcanic Belt in Mexico
-(Puebla, north of Guerrero State and Jalisco.\\
-
-\begin{figure}
-\centering
-\includegraphics{all_figures/figure_4.png}
-\caption{Suitability prevalence area. Here we observe the harmonic mean
-of suitability for each pixel along the distribucion from 0 to 1. The
-bold lines shows the the area when the harmonic mean es above 0.5
-threshold (blud line) and 0.9 threshold (black line)}
-\end{figure}
-
-\hypertarget{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
-\subsubsection{Fst as a funciton of SPA and Fixation index projected in
-geography}\label{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}
-
-For all the linear regressions performed (Table 1), the only one that
-had statistical significance is the one for the Sciurus aberti barberi
-population with an R-squared value of 0.794.
-
-\begin{longtable}[]{@{}ccc@{}}
-\caption{R squared from linear models (fixation index as a function of
-suitability prevalence area). The only R squared with statistical
-significance is barberi}\tabularnewline
-\toprule\noalign{}
-population & r.squared & p.value \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-population & r.squared & p.value \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-S. a. aberti Carson-SFW & 0.000 & 0.962 \\
-S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
-S. a. aberti MT-Zuni & 0.002 & 0.894 \\
-S. a. aberti San Juan & 0.000 & 0.952 \\
-S. a. barberi & 0.794 & 0.001 \\
-S. a. chuscensis E & 0.002 & 0.892 \\
-S. a. chuscensis W & 0.018 & 0.713 \\
-S. a. ferreus Carson E & 0.040 & 0.581 \\
-S. a. ferreus Pike & 0.000 & 0.967 \\
-S. a. kaibabensis & 0.005 & 0.841 \\
-\end{longtable}
-
-It can be noted that S. a. aberti Coconino-Gila had an R-squared value
-greater than 0 (0.131) but it is not statistically significant.
-Furthermore, in our approach we imputed the value of the origin with an
-Fst of zero. To show that the model hypothesis still holds, we performed
-a regression for barberi without the point of origin imposed (Figure
-5b), in addition to eliminating the S. aberti chuscensis populations due
-to their low uncertainty in georeference, obtaining an R-squared of 0.72
-(p value = 0.016).
-
-\begin{figure}
-\includegraphics[width=0.5\linewidth]{all_figures/figure_5a} \includegraphics[width=0.5\linewidth]{all_figures/figure_5b} \caption{Linear regression for S a.
-barberi population A) SPA in all populations B) withouth barberi and chusensis populations.
-We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)}\label{fig:figures-side}
-\end{figure}
-
-From the model we extrapolated the values from SPA along the geography
-to new \(F_{st}\) values to get a geographical projection of fixation
-index from \emph{S. a. barberi} population. (Figure 6)
-
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth,height=\textheight]{all_figures/figure_6.png}
-\caption{Projection of the \(F_{st}\) model to the geography in the
-actual distribution of \emph{S aberti}}
-\end{figure}
-
-\hypertarget{references}{%
-\subsubsection*{References}\label{references}}
-\addcontentsline{toc}{subsubsection}{References}
-
 \hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-beyer2020high}{}}%
-Beyer, Robert M, Mario Krapp, and Andrea Manica. 2020.
-{``High-Resolution Terrestrial Climate, Bioclimate and Vegetation for
-the Last 120,000 Years.''} \emph{Scientific Data} 7 (1): 236.
-
 \leavevmode\vadjust pre{\hypertarget{ref-bono2018genome}{}}%
@@
 Mammalogy}. Oxford University Press US.
-
-\leavevmode\vadjust pre{\hypertarget{ref-guisan2005predicting}{}}%
-Guisan, Antoine, and Wilfried Thuiller. 2005. {``Predicting Species
-Distribution: Offering More Than Simple Habitat Models.''} \emph{Ecology
-Letters} 8 (9): 993--1009.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hedrick2005standardized}{}}%
-Hedrick, Philip W. 2005. {``A Standardized Genetic Differentiation
-Measure.''} \emph{Evolution} 59 (8): 1633--38.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karlin1968rates}{}}%
-Karlin, Samuel. 1968. {``Rates of Approach to Homozygosity for Finite
-Stochastic Models with Variable Population Size.''} \emph{The American
-Naturalist} 102 (927): 443--55.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-merow2013practical}{}}%
-Merow, Cory, Matthew J Smith, and John A Silander Jr. 2013. {``A
-Practical Guide to MaxEnt for Modeling Species' Distributions: What It
-Does, and Why Inputs and Settings Matter.''} \emph{Ecography} 36 (10):
-1058--69.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nei1986definition}{}}%
-Nei, Masatoshi. 1986. {``Definition and Estimation of Fixation
-Indices.''} \emph{Evolution} 40 (3): 643--45.
-
 \leavevmode\vadjust pre{\hypertarget{ref-nogues2009predicting}{}}%
@@
 (5): 521--31.
-
-\leavevmode\vadjust pre{\hypertarget{ref-owens2013constraints}{}}%
-Owens, Hannah L, Lindsay P Campbell, L Lynnette Dornak, Erin E Saupe,
-Narayani Barve, Jorge Soberón, Kate Ingenloff, et al. 2013.
-{``Constraints on Interpretation of Ecological Niche Models by Limited
-Environmental Ranges on Calibration Areas.''} \emph{Ecological
-Modelling} 263: 10--18.
-
-\leavevmode\vadjust pre{\hypertarget{ref-peterson2008rethinking}{}}%
-Peterson, A Townsend, Monica Papeş, and Jorge Soberón. 2008.
-{``Rethinking Receiver Operating Characteristic Analysis Applications in
-Ecological Niche Modeling.''} \emph{Ecological Modelling} 213 (1):
-63--72.
-
-\leavevmode\vadjust pre{\hypertarget{ref-peterson2012species}{}}%
-Peterson, A Townsend, and Jorge Soberón. 2012. {``Species Distribution
-Modeling and Ecological Niche Modeling: Getting the Concepts Right.''}
-\emph{Natureza \& Conserva{ç}{ã}o} 10 (2): 102--7.
-
-\leavevmode\vadjust pre{\hypertarget{ref-phillips2006maximum}{}}%
-Phillips, Steven J, Robert P Anderson, and Robert E Schapire. 2006.
-{``Maximum Entropy Modeling of Species Geographic Distributions.''}
-\emph{Ecological Modelling} 190 (3-4): 231--59.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saupe2012variation}{}}%
-Saupe, EE, V Barve, CE Myers, J Soberón, N Barve, CM Hensz, AT Peterson,
-H Lc Owens, and A Lira-Noriega. 2012. {``Variation in Niche and
-Distribution Model Performance: The Need for a Priori Assessment of Key
-Causal Factors.''} \emph{Ecological Modelling} 237: 11--22.
-
-\leavevmode\vadjust pre{\hypertarget{ref-singarayer2010high}{}}%
-Singarayer, Joy S, and Paul J Valdes. 2010. {``High-Latitude Climate
-Sensitivity to Ice-Sheet Forcing over the Last 120 Kyr.''}
-\emph{Quaternary Science Reviews} 29 (1-2): 43--55.
-
-\leavevmode\vadjust pre{\hypertarget{ref-soberon2005interpretation}{}}%
-Soberon, Jorge, and A Townsend Peterson. 2005. {``Interpretation of
-Models of Fundamental Ecological Niches and Species' Distributional
-Areas.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-soberon2009niches}{}}%
-Soberón, Jorge, and Miguel Nakamura. 2009. {``Niches and Distributional
-Areas: Concepts, Methods, and Assumptions.''} \emph{Proceedings of the
-National Academy of Sciences} 106 (supplement\_2): 19644--50.
-
-\leavevmode\vadjust pre{\hypertarget{ref-soberon2020shape}{}}%
-Soberón, Jorge, and A Townsend Peterson. 2020. {``What Is the Shape of
-the Fundamental Grinnellian Niche?''} \emph{Theoretical Ecology} 13 (1):
-105--15.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-upham2019inferring}{}}%
-Upham, Nathan S, Jacob A Esselstyn, and Walter Jetz. 2019. {``Inferring
-the Mammal Tree: Species-Level Sets of Phylogenies for Questions in
-Ecology, Evolution, and Conservation.''} \emph{PLoS Biology} 17 (12):
-e3000494.
-
-\leavevmode\vadjust pre{\hypertarget{ref-valdes2017bridge}{}}%
-Valdes, Paul J, Edward Armstrong, Marcus PS Badger, Catherine D
-Bradshaw, Fran Bragg, Michel Crucifix, Taraka Davies-Barnard, et al.
-2017. {``The BRIDGE HadCM3 Family of Climate Models: HadCM3@ Bristol V1.
-0.''} \emph{Geoscientific Model Development} 10 (10): 3715--43.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%
-Wright, Sewall. 1943. {``Isolation by Distance.''} \emph{Genetics} 28
-(2): 114.
-
 \end{CSLReferences}

### 2023-07-29 manuscript_english.Rmd
changes in manuscript

--- 2023-07-10 manuscript_english.Rmd
@@
 
-For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$.
-We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$.
-Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions.
-If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$.
-i.
-e.
-the set of environmental conditions for which values of fitness are capable of sustaining populations.
-We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$.
+We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables.
+Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
+that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.
+
+That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
+If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
+
 This is a way of expressing the fundamental niche as a function of environmental conditions.
 So we can map each point of geography to ambient space.
-and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
-However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$.
+Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
+However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates.
+However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.
+
 Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.
+Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
+$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.
+
+We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space.
+Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.
+
+![Hypothetical niche in environmental space.
+An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
 
@@
 This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$.
-In a practical way, instead of $\mathbf{N}_F$, we approximated
-$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
-(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.
+That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$.
+In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.
 
-The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
+Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly.
+in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
 
-Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
+The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
+
+Es decir:
+
 $$
@@
 $$
-
-![Hypothetical niche in environmental space.
-An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}
-
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates.
-So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:
-
-$$
-N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
-$$
+Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
 
@@
 
-1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
+1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
 2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
-3. Characterize past environments and retain projections only for sites with analogous environmental conditions to the present.
-4. Calculate the harmonic mean of the suitability for all the sites all time slices.
+3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
+4. Generate a harmonic mean suitability map for all time slices.
 
@@
 For each scenario into the past, 10 replicates were cross-validated and averaged.
+For each projection into the past, as well as the model of the present, we generated a binary map at a suitability threshold of 90\% to get an estimate of the area distribution for each time slice in pixels.
+In addition we fit a locally estimated scatterplot smoothing (loess) model to the time series trend area using the R package stats in order to observe patterns in the area - timeline scatterplot.
 
@@
 ### Endemic historic Area
-To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $\mathbf{S}$ equation.
+To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $mathbf{S}$ equation.
 This threshold is the proportion of time that pixel has had ideal conditions.
@@
 ### Current and past distribution
-We show the species distribution model for *Sciurus aberti*  along the time line of the projected area per time slice. We show 5 notable moments in time line. 
-and 5 notable points 
+We get the binary maps for all the models in suitability threshold of 0.9 and extract the number of pixels above this threshold in order to get a relative predicted area.
+Here we show the trend from present (Figure 3A) to past of the the projected area of *Sciurus aberti* along the time line (Figure 3F).
+The trend pattern exhibits an inverted U-shape, showing that the present area is similar to the 120 000 area, with a maximum in 22 000 coinciding with the Last Glacial Maximum in North America (Figure 3B).
+We highlight three other notable scenarios, the first local minimum after the interglacial maximum appears in the year 30 000 (Figure 3C) coinciding with a distribution similar to the present but with more suitable regions emerging in New Mexico and Colorado; a local maximum in year 62 000 at the middle of the trend line (Figure 3D), with a more suitable potential distribution towards the south, with conditions disappearing practically everywhere in the United States, except in the border with Sonora, but connecting the area of the Sierra Madre Occidental with Zacatecas, San Luis Potosí and the appearance of a corridor over the Transverse Volcanic Axis (from Jalisco to Tlaxcala); finally we show a minimum area of the whole time series in year 112 000.
+(Figure 3E), with the disappearance of conditions in practically all the places where there were expansions, preserving conditions only in some places of Chihuahua (Campo Verde and Tutuaca Natural Protected Areas), Durango and Arizona (Prescott and Sedona).
+
 <p align="center">
@@
 ### SPA and Endemic Historic Area
-We get the suitability presence area (figure 2) delmited by a 0.5 and 0.9 threshold (Figure 4).
-The endemic region where the enviornmental conditions 90 \% del tiempo is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango.
+We get the suitability prevalence area delimited by a 0.5 and 0.9 threshold (Figure 4).
+The endemic region where the enviornmental conditions 90 \% of the time is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango.
 We correspond this Area to the *Sciurus aberti barbieri* subspecies.
@@
 
-### Fst as a funciton of SPA and Fixation index projected in geography
+### $F_{st}$ as a funciton of SPA and Fixation index projected in geography
 
@@
 We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)", echo=FALSE}
+
 knitr::include_graphics(c("all_figures/figure_5a.png","all_figures/figure_5b.png"))
@@
 
-From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S. a. barberi* population. (Figure 6)
+From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S.
+a.
+barberi* population.
+(Figure 6).
+The lowest values are close to the Sierra Madre Occidental Area, in the limits of Chihuahua and Sonora States as well as Sinaloa and Durango, score higher near to the Tutuaca Wildlife regue There is also a hotspot between Sonora And arizona, near to Bisbee Arizona in United States as well as Naco Sonora y Mexico.
 
@@
 
+## Discussion
+
+With the current availability of occurrence data and models of potential distribution projected to past climate scenarios it is possible to find geographic patterns of differentiation in populations at an acceptable resolution (half a degree) that allows us to define sites where the species can potentially be found but share different degrees of genetic diversity.
+
+Projections of genetic diversity in geography have been made in previous studies @van2012mapping.
+However in a general way they are extrapolations of observations, without a predictive model, which with some explanatory variable behind can explain in a statistically convincing way the information obtained from population genetics.
+
+We have shown a generalizable protocol for predicting patterns of population genetics in geography.
+In this way we link information on the evolutionary history of a taxon and its ecology.
+In addition, we provide a predictive vehicle that generates new geographic information applicable to decision making in both survey conduct and taxon conservation.
+
+Thanks to the theoretical support that we have briefly outlined in this paper, it is possible to explain the explanatory value of the Coefficient of determination for our linear model ($R^2 = 0.79 $), which is information coming from two completely different sources, without any correlation, but always guided by the premise of the evolutionary process as causal.
+
+These maps represent a turning point in the conception of conservation.
+With the tool we propose in this study, it is possible to delimit conservation sites based on evidence of current and historical distribution and genetic diversity among different populations.
+
+While many conservation decisions are made on the basis of genetic diversity, studying the genetic structure of populations at a macro-scale is often very complicated, and making conservation decisions based on current distribution without taking into account the historical distribution of taxa can lead to a misguided conservation strategy.
+
+The generalization of our protocol and its implementation in tools widely used by biogeography allow its direct implementation in many case studies with available information, as well as the design of new questions, e.g.
+species assemblages under prevalence of environmental suitability.
+A research path was also established on whether the process of environmental stability over time is related to biodiversity.
+
 ### References

### 2023-07-29 manuscript_english.tex
changes in manuscript

--- 2023-07-11 manuscript_english.tex
@@
 
+As a case study we consider squirrels (\emph{Sciurus aberti}), a species
+currently distributed in disjunct patches from the southern Rocky
+Mountains in the United States to the northern Sierra Madre Occidental
+in Mexico for which it is possible to find reliable information on both
+the genetic structure of its populations and its current distribution
+(Bono et al. 2018; Burgin et al. 2018).
+
+Our results reveal that fitness prevalence corresponds to the fixation
+index of \emph{S. aberti} populations with respect to a source
+population. Populations closer to the historical endemic area present a
+higher genetic diversity and a lower \(F_{st}\) value. Finally, this
+study allows us to add a biogeographic explanation to the results
+obtained with population genetic methods and to generate maps of this
+structure as tools to support conservation with a perspective that
+integrates both population genetics and historical patterns of species
+distribution.
+
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
+
 The current distributions of species do not always correspond to their
@@
 
-As a case study we consider squirrels (\emph{Sciurus aberti}), a species
-currently distributed in disjunct patches from the southern Rocky
-Mountains in the United States to the northern Sierra Madre Occidental
-in Mexico for which it is possible to find reliable information on both
-the genetic structure of its populations and its current distribution
-(Bono et al. 2018; Burgin et al. 2018).
-
-Our results reveal that suitability prevalence corresponds to the
-fixation index of \emph{S. aberti} populations with respect to a source
-population. Populations closer to the historical endemic area present a
-higher genetic diversity and a lower \(F_{st}\) value. Finally, this
-study allows us to add a biogeographic explanation to the results
-obtained with population genetic methods and to generate maps of this
-structure as tools to support conservation with a perspective that
-integrates both population genetics and historical patterns of species
-distribution.
-
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
 \hypertarget{theoretical-basis}{%
@@
 
-For a given region \(\mathbf{G}\) in te geography, at any point in time,
-we have a set of existing environmental conditions
-\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
-We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
+We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
 environmental variables, that is, a point in the space of \(v\)
-environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
-\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
-that is, for a species with non-overlapping generations, the
-population's net reproductive rate \(R_0\) is a measure of fitness that
-depends on environmental conditions. If we take the set of places in
-environmental space where these values are greater than a threshold, we
-can define a volume in multidimensional space such that \(R(\vec{e})\)
-is always positive and greater than this threshold, that is
-\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
-environmental conditions for which values of fitness are capable of
-sustaining populations. We call
-\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
-expressing the fundamental niche as a function of environmental
-conditions. So we can correspong each point of geography to
-environmental space, with different environmental conditions and
-different growth rates. However all sites within the region defined by
-\(\mathbf{N}_F\) will have viable populations and positive rates
-depending on time because we assume that the environment is dynamic and
-each geographic point is static, so for each site in geographic space
-this growth rate depends on the environment and therefore is different
-at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
-will be related to the local population size as long as it is within the
-boundary defined by \(\mathbf{N}_F\).
+environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
+relates fitness to environment, that is, for a species with
+non-overlapping generations, the population's net reproductive rate
+\(R_i\) is a measure of fitness that depends on environmental
+conditions.
+
+That is, the sites in the geography \(\mathbf{G}_o\) that present
+positive growth rates are related to favorable conditions. If we take
+the set of places in ambient space where these values are greater than a
+threshold, we can define a volume in multidimensional space such that
+\(R(\vec{e})\) is always positive and greater than this threshold, that
+is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
+
+This is a way of expressing the fundamental niche as a function of
+environmental conditions. So we can map each point of geography to
+ambient space. Thus, for each site in the geography there will
+correspond a site in the environmental space with different
+environmental conditions and different growth rates. However all sites
+within the region defined by \(\mathbf{N}_F\) will have viable
+populations and positive rates. However, we assume that the environment
+is dynamic and each geographic point is static, so for each site in
+geographic space this growth rate depends on the environment and
+therefore is different at each time.
+
+Thus locally the growth rate will be related to the local population
+size as long as it is within the boundary defined by \(\mathbf{N}_F\).
+Therefore, the local population size as a function of the intrinsic rate
+that depends on the change in environmental conditions over time is:
+\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
+population size is a function of ideal environmental conditions.
+
+We start by considering this defined volume of the set of \(R_0\) as a
+set of trajectories for each point from geographic space to
+environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
+the population grows, that is,
+\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
+any local population will have a positive growth rate.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
+\caption{Hypothetical niche in environmental space. An arbitrary point
+moves within the ambient space entering and exiting the volume (niche)
+where the intrinsic growth rate is positive.}
+\end{figure}
 
@@
 That is, the fitness-related environmental suitability can be any
-function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
-\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
-values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
-practical way, instead of \(\mathbf{N}_F\), we approximated
-\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
-existing niche) from known presence points in the geography
-\(\mathbf{G}\) obtained from emprical observations stored in database,
-and map to \(\mathbf{E}\) in order to numerically fitting a convex
-volume of the environmental space that enclose the known information.
-
-The objective of our approach is not to know the function of \(R_0\) but
-rather to relate this local variation and the function \(\mathbf{N}_F\)
-with a function \(S_i(\vec{e}_i)\) of environmental suitability that can
-be estimated from known methods of applied mathematics and machine
-learning (for example the case of Maxent, Phillips, Anderson, and
-Schapire (2006)).
-
-Population size locally is a function of environmental suitability,
-where if suitability is \(0\) the population is \(0\) and if suitability
-is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) \[
-N_i \propto  S_i(\vec{e}_i)
-\]
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
-\caption{Hypothetical niche in environmental space. An arbitrary point
-moves within the environmental space entering and exiting the volume
-(niche) where the intrinsic growth rate is positive.}
-\end{figure}
+function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
+\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
+of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
+approximated from points of presence in the geography obtained from
+databases, numerically approximating a convex surface with a surface
+\(\mathbf{N}_R\) of the environmental space.
 
@@
 point in the geography is static and has a trajectory in the
-environmental space and for each trajectory the fitness function changes
-and therefore the suitability function changes but not in a homogeneous
-way, so each site presents different displacement rates in envrionmental
-spaces and as a consequence, different demographic rates. So we can
-establish that for each site \(i\) in geographic space \(\mathbf{G}\),
-there is a population size \(N_i\) that changes with respect to a
-fitness function dependent on time-varying environmental conditions:
+environmental space and that for each trajectory the fitness function
+changes and therefore the suitability function changes but not
+accordingly. in a homogeneous way, if not more so that each site
+presents different displacement rates and different demographic rates.
+
+The objective of this theoretical approach is not to know the shape of
+the function of \(R_0\) but rather to find this local variation relating
+this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
+\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
+estimated from known methods of applied mathematics and machine learning
+(for example the case of Maxent, Phillips, Anderson, and Schapire
+(2006)).
+
+Es decir:
 
 \[
-N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
-\]
+N_i \propto  S_i(\vec{e}_i)
+\] Population size locally is a function of environmental suitability,
+where if suitability is \(0\) the population is \(0\) and if suitability
+is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)
 
@@
 \(\tau\) generations
-\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\)
-for each \(i\) site in \(\mathbf{G}\). We substitute the harmonic mean
-into the above equation, considering that although the number of
-generations is not exactly equal to the time of the fixation index, they
-are expected to be very close:
+\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
+We substitute the harmonic mean into the above equation, considering
+that although the number of generations is not exactly equal to the time
+of the fixation index, they are expected to be very close.
 
 \[
-F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
+F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
 \]
@@
 loss of heterozygosity (through the fixation index \(F_{st}\)) is
-inversely proportional to the suitability prevalence for an \(i\) site
-in the geography:
+inversely proportional to the fitness prevalence for an \(i\) site in
+the geography:
 
@@
 \] If we consider all sites in the available geography, we call this
-harmonic \emph{Suitability Prevalence Area}
+harmonic mean of fitness \emph{Suitability prevalence Area}
 \(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
@@
 \item
-  Model the potential distribution of a taxon with present-day
+  model the potential distribution of a taxon with present-day
   environmental conditions and obtain an environmental suitability map.
@@
 \item
-  Characterize past environments and retain projections only for sites
-  with analogous environmental conditions to the present.
+  Characterize past analogous climates and retain projections only for
+  sites with climates analogous to the present.
 \item
-  Calculate the harmonic mean of the suitability for all the sites all
-  time slices.
+  Generate a harmonic mean suitability map for all time slices.
 \end{enumerate}
@@
 000 years. For each scenario into the past, 10 replicates were
-cross-validated and averaged.
+cross-validated and averaged. For each projection into the past, as well
+as the model of the present, we generated a binary map at a suitability
+threshold of 90\% to get an estimate of the area distribution for each
+time slice in pixels. In addition we fit a locally estimated scatterplot
+smoothing (loess) model to the time series trend area using the R
+package stats in order to observe patterns in the area - timeline
+scatterplot.
 
@@
 threshold is applied to generate a binary map according to the
-\(\mathbf{S}\) equation. This threshold is the proportion of time that
+\(mathbf{S}\) equation. This threshold is the proportion of time that
 pixel has had ideal conditions. For our study, we consider the value of
@@
 (2018). It consideres 10 populations along the known distribution and
-provides values of \(F_{st}\) between this populations in matrix form.
-The matrix was transformed into a table as pairs between populations
-with index values between each pair. The values of the population with
+provides values of \(F_st\) between this populations in matrix form. The
+matrix was transformed into a table as pairs between populations with
+index values between each pair. The values of the population with
 itself, i.e.~\(0\), were included. Subsequently, georeferences were
@@
 
-We show the species distribution model for \emph{Sciurus aberti} along
-the time line of the projected area per time slice. We show 5 notable
-moments in time line. and 5 notable points
+We get the binary maps for all the models in suitability threshold of
+0.9 and extract the number of pixels above this threshold in order to
+get a relative predicted area. Here we show the trend from present
+(Figure 3A) to past of the the projected area of \emph{Sciurus aberti}
+along the time line (Figure 3F). The trend pattern exhibits an inverted
+U-shape, showing that the present area is similar to the 120 000 area,
+with a maximum in 22 000 coinciding with the Last Glacial Maximum in
+North America (Figure 3B). We highlight three other notable scenarios,
+the first local minimum after the interglacial maximum appears in the
+year 30 000 (Figure 3C) coinciding with a distribution similar to the
+present but with more suitable regions emerging in New Mexico and
+Colorado; a local maximum in year 62 000 at the middle of the trend line
+(Figure 3D), with a more suitable potential distribution towards the
+south, with conditions disappearing practically everywhere in the United
+States, except in the border with Sonora, but connecting the area of the
+Sierra Madre Occidental with Zacatecas, San Luis Potosí and the
+appearance of a corridor over the Transverse Volcanic Axis (from Jalisco
+to Tlaxcala); finally we show a minimum area of the whole time series in
+year 112 000. (Figure 3E), with the disappearance of conditions in
+practically all the places where there were expansions, preserving
+conditions only in some places of Chihuahua (Campo Verde and Tutuaca
+Natural Protected Areas), Durango and Arizona (Prescott and Sedona).
 
@@
 
-We get the suitability presence area (figure 2) delmited by a 0.5 and
-0.9 threshold (Figure 4). The endemic region where the enviornmental
-conditions 90 \% del tiempo is the Sierra Madre Occidental, in the
+We get the suitability prevalence area delimited by a 0.5 and 0.9
+threshold (Figure 4). The endemic region where the enviornmental
+conditions 90 \% of the time is the Sierra Madre Occidental, in the
 limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this
@@
 
-\hypertarget{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
-\subsubsection{Fst as a funciton of SPA and Fixation index projected in
-geography}\label{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}
+\hypertarget{f_st-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
+\subsubsection{\texorpdfstring{\(F_{st}\) as a funciton of SPA and
+Fixation index projected in
+geography}{F\_\{st\} as a funciton of SPA and Fixation index projected in geography}}\label{f_st-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}
 
@@
 to new \(F_{st}\) values to get a geographical projection of fixation
-index from \emph{S. a. barberi} population. (Figure 6)
+index from \emph{S. a. barberi} population. (Figure 6). The lowest
+values are close to the Sierra Madre Occidental Area, in the limits of
+Chihuahua and Sonora States as well as Sinaloa and Durango, score higher
+near to the Tutuaca Wildlife regue There is also a hotspot between
+Sonora And arizona, near to Bisbee Arizona in United States as well as
+Naco Sonora y Mexico.
 
@@
 \end{figure}
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
+
+With the current availability of occurrence data and models of potential
+distribution projected to past climate scenarios it is possible to find
+geographic patterns of differentiation in populations at an acceptable
+resolution (half a degree) that allows us to define sites where the
+species can potentially be found but share different degrees of genetic
+diversity.
+
+Projections of genetic diversity in geography have been made in previous
+studies Van Zonneveld et al. (2012). However in a general way they are
+extrapolations of observations, without a predictive model, which with
+some explanatory variable behind can explain in a statistically
+convincing way the information obtained from population genetics.
+
+We have shown a generalizable protocol for predicting patterns of
+population genetics in geography. In this way we link information on the
+evolutionary history of a taxon and its ecology. In addition, we provide
+a predictive vehicle that generates new geographic information
+applicable to decision making in both survey conduct and taxon
+conservation.
+
+Thanks to the theoretical support that we have briefly outlined in this
+paper, it is possible to explain the explanatory value of the
+Coefficient of determination for our linear model (\$R\^{}2 = 0.79 \$),
+which is information coming from two completely different sources,
+without any correlation, but always guided by the premise of the
+evolutionary process as causal.
+
+These maps represent a turning point in the conception of conservation.
+With the tool we propose in this study, it is possible to delimit
+conservation sites based on evidence of current and historical
+distribution and genetic diversity among different populations.
+
+While many conservation decisions are made on the basis of genetic
+diversity, studying the genetic structure of populations at a
+macro-scale is often very complicated, and making conservation decisions
+based on current distribution without taking into account the historical
+distribution of taxa can lead to a misguided conservation strategy.
+
+The generalization of our protocol and its implementation in tools
+widely used by biogeography allow its direct implementation in many case
+studies with available information, as well as the design of new
+questions, e.g.~species assemblages under prevalence of environmental
+suitability. A research path was also established on whether the process
+of environmental stability over time is related to biodiversity.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-van2012mapping}{}}%
+Van Zonneveld, Maarten, Xavier Scheldeman, Pilar Escribano, Maria A
+Viruel, Patrick Van Damme, Willman Garcia, César Tapia, José Romero,
+Manuel Siguenas, and José I Hormaza. 2012. {``Mapping Genetic Diversity
+of Cherimoya (Annona Cherimola Mill.): Application of Spatial Analysis
+for Conservation and Use of Plant Genetic Resources.''} \emph{PloS One}
+7 (1): e29845.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%

### 2023-07-29 manuscript_english.Rmd
finish abstract

@@
 ## Abstract
-As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
 
-Our results reveal that fitness prevalence corresponds to the fixation index of *S.
-aberti* populations with respect to a source population.
+Genetic variation among populations exhibits biogeographic patterns.
+In this paper we establish a theoretical framework based on environmental Grinellian niches as a tool for modeling species distributions, as well as their projected past environmental scenarios to relate the present evolutionary history in both the geographic distribution and genetic structure of their populations.
+Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework with the help of the joint analysis of spatial, environmental and genetic data, with tools that are widely distributed among biogeographers and evolutionary biologists.
+Our results reveal that when we have case of isolation by distance, suitability prevalence along time corresponds to the fixation index of populations with respect to a source population.
 Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value.
-Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
+This framework allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
 
@@
 We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables.
-Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
-that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.
+Then $R_i(\vec{e}_i)$ is a function that relates evolutionary fitness to environment,
+that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of evolutionary fitness that depends on environmental conditions.
 
@@
 
-This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of fitness and environment.
+This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of evolutionary fitness and environment.
 In that study they tested the validity with monthly variables and empirical values on a short time scale.
@@
 
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly.
-in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
+Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory, the fitness function changes and, therefore, the suitability function changes but not accordingly, in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
 
-The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).
-
-Es decir:
+The goal of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum), i.e.
+:
 
@@
 $$
-If we consider all sites in the available geography, we call this harmonic mean of fitness *Suitability prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
+If we consider all sites in the available geography, we call this harmonic mean *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ (SPA) since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
 environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred.
-Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.
+Conversely, it is expected that in populations with a high fixation index value, suitability will not have prevailed over the same time interval even though conditions currently exist and populations are present.
 
@@
 
-To test our model we use the Aberti squirrel (*Sciurus aberti*) as a case study because:
+As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
+In order to select this species, we list the following considerations:
 
@@
 - The subspecies are geographically isolated, since they are found in pine forests surrounded by desert. Therefore we can consider the assumption of isolation by distance [@wright1943isolation]. 
-- There are sufficient occurrence records for the entire range of the known distribution.
 - It has low speciation rate (0.25) [@upham2019inferring], therefore, we assume that its fundamental niche is stable over time.
@@
 To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered.
-For this work we considerer the data from @@bono2018genome.
+For this work we considerer the data from @bono2018genome.
 It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form.
@@
 
-For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an R-squared value of 0.794.
+For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an $R^2$ value of $0.794$.
 
-```{r, echo=FALSE}
-aberti_pops <- read.csv(file = "all_tables/fst_SPA_rsquared.csv")
-
-knitr::kable(aberti_pops, "simple", align = "ccc", caption = "R squared from linear models (fixation index as a function of suitability prevalence area).
-The only R squared with statistical significance is barberi")
-```
+|          Population          | R squared | p value |
+|:----------------------------:|:---------:|:-------:|
+|   _S. a. aberti Carson-SFW_  |     0     |  0.962  |
+| _S. a. aberti Coconino-Gila_ |   0.131   |  0.304  |
+|    _S. a. aberti MT-Zuni_    |   0.002   |  0.894  |
+|    _S. a. aberti San Juan_   |     0     |  0.952  |
+|        _S. a. barberi_       |   0.794   |  0.001  |
+|     _S. a. chuscensis E_     |   0.002   |  0.892  |
+|     _S. a. chuscensis W_     |   0.018   |  0.713  |
+|   _S. a. ferreus Carson E_   |    0.04   |  0.581  |
+|     _S. a. ferreus Pike_     |     0     |  0.967  |
+|      _S. a. kaibabensis_     |   0.005   |  0.841  |
 

### 2023-07-29 manuscript_english.tex
finish abstract

@@
 
-As a case study we consider squirrels (\emph{Sciurus aberti}), a species
-currently distributed in disjunct patches from the southern Rocky
-Mountains in the United States to the northern Sierra Madre Occidental
-in Mexico for which it is possible to find reliable information on both
-the genetic structure of its populations and its current distribution
-(Bono et al. 2018; Burgin et al. 2018).
-
-Our results reveal that fitness prevalence corresponds to the fixation
-index of \emph{S. aberti} populations with respect to a source
-population. Populations closer to the historical endemic area present a
-higher genetic diversity and a lower \(F_{st}\) value. Finally, this
-study allows us to add a biogeographic explanation to the results
-obtained with population genetic methods and to generate maps of this
-structure as tools to support conservation with a perspective that
-integrates both population genetics and historical patterns of species
-distribution.
+Genetic variation among populations exhibits biogeographic patterns. In
+this paper we establish a theoretical framework based on environmental
+Grinellian niches as a tool for modeling species distributions, as well
+as their projected past environmental scenarios to relate the present
+evolutionary history in both the geographic distribution and genetic
+structure of their populations. Thanks to the advance of next-generation
+sequencing technologies, as well as more accurate climate models and an
+enormous amount of information stored in biological collections, it is
+possible to implement this theoretical framework with the help of the
+joint analysis of spatial, environmental and genetic data, with tools
+that are widely distributed among biogeographers and evolutionary
+biologists. Our results reveal that when we have case of isolation by
+distance, suitability prevalence along time corresponds to the fixation
+index of populations with respect to a source population. Populations
+closer to the historical endemic area present a higher genetic diversity
+and a lower \(F_{st}\) value. This framework allows us to add a
+biogeographic explanation to the results obtained with population
+genetic methods and to generate maps of this structure as tools to
+support conservation with a perspective that integrates both population
+genetics and historical patterns of species distribution.
 
@@
 environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
-relates fitness to environment, that is, for a species with
+relates evolutionary fitness to environment, that is, for a species with
 non-overlapping generations, the population's net reproductive rate
-\(R_i\) is a measure of fitness that depends on environmental
-conditions.
+\(R_i\) is a measure of evolutionary fitness that depends on
+environmental conditions.
 
@@
 This equation was proposed by Soberon and Peterson Soberón and Peterson
-(2020) to relate demographics to a function of fitness and environment.
-In that study they tested the validity with monthly variables and
-empirical values on a short time scale. relating life history parameters
-and the environmental suitability of niche models.
+(2020) to relate demographics to a function of evolutionary fitness and
+environment. In that study they tested the validity with monthly
+variables and empirical values on a short time scale. relating life
+history parameters and the environmental suitability of niche models.
 
@@
 point in the geography is static and has a trajectory in the
-environmental space and that for each trajectory the fitness function
-changes and therefore the suitability function changes but not
-accordingly. in a homogeneous way, if not more so that each site
+environmental space and that for each trajectory, the fitness function
+changes and, therefore, the suitability function changes but not
+accordingly, in a homogeneous way, if not more so that each site
 presents different displacement rates and different demographic rates.
 
-The objective of this theoretical approach is not to know the shape of
-the function of \(R_0\) but rather to find this local variation relating
+The goal of this theoretical approach is not to know the shape of the
+function of \(R_0\) but rather to find this local variation relating
 this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
@@
 (for example the case of Maxent, Phillips, Anderson, and Schapire
-(2006)).
-
-Es decir:
+(2006)), i.e.~:
 
@@
 \] If we consider all sites in the available geography, we call this
-harmonic mean of fitness \emph{Suitability prevalence Area}
-\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
-\(F_{st}\) is lower are in areas where the prevalence over time is
-higher, i.e.~environmental conditions remain suitable for a significant
-period of time and therefore no effects of genetic drift of populations
-have occurred. Conversely, it is expected that in populations with a
-high fixation index value, fitness will not have prevailed over the same
-time interval even though conditions currently exist and populations are
-present.
+harmonic mean \emph{Suitability Prevalence Area}
+\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
+where the \(F_{st}\) is lower are in areas where the prevalence over
+time is higher, i.e.~environmental conditions remain suitable for a
+significant period of time and therefore no effects of genetic drift of
+populations have occurred. Conversely, it is expected that in
+populations with a high fixation index value, suitability will not have
+prevailed over the same time interval even though conditions currently
+exist and populations are present.
 
@@
 
-To test our model we use the Aberti squirrel (\emph{Sciurus aberti}) as
-a case study because:
+As a case study we consider squirrels (\emph{Sciurus aberti}), a species
+currently distributed in disjunct patches from the southern Rocky
+Mountains in the United States to the northern Sierra Madre Occidental
+in Mexico for which it is possible to find reliable information on both
+the genetic structure of its populations and its current distribution
+(Bono et al. 2018; Burgin et al. 2018). In order to select this species,
+we list the following considerations:
 
@@
   assumption of isolation by distance (Wright 1943).
-\item
-  There are sufficient occurrence records for the entire range of the
-  known distribution.
 \item
@@
 had statistical significance is the one for the Sciurus aberti barberi
-population with an R-squared value of 0.794.
+population with an \(R^2\) value of \(0.794\).
 
 \begin{longtable}[]{@{}ccc@{}}
-\caption{R squared from linear models (fixation index as a function of
-suitability prevalence area). The only R squared with statistical
-significance is barberi}\tabularnewline
 \toprule\noalign{}
-population & r.squared & p.value \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-population & r.squared & p.value \\
+Population & R squared & p value \\
 \midrule\noalign{}
@@
 \endlastfoot
-S. a. aberti Carson-SFW & 0.000 & 0.962 \\
-S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
-S. a. aberti MT-Zuni & 0.002 & 0.894 \\
-S. a. aberti San Juan & 0.000 & 0.952 \\
-S. a. barberi & 0.794 & 0.001 \\
-S. a. chuscensis E & 0.002 & 0.892 \\
-S. a. chuscensis W & 0.018 & 0.713 \\
-S. a. ferreus Carson E & 0.040 & 0.581 \\
-S. a. ferreus Pike & 0.000 & 0.967 \\
-S. a. kaibabensis & 0.005 & 0.841 \\
+\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
+\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
+\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
+\emph{S. a. aberti San Juan} & 0 & 0.952 \\
+\emph{S. a. barberi} & 0.794 & 0.001 \\
+\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
+\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
+\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
+\emph{S. a. ferreus Pike} & 0 & 0.967 \\
+\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
 \end{longtable}

### 2023-08-01 manuscript_english.tex
Edits to the Intro

@@
 
-Genetic variation among populations exhibits biogeographic patterns. In
-this paper we establish a theoretical framework based on environmental
-Grinellian niches as a tool for modeling species distributions, as well
-as their projected past environmental scenarios to relate the present
-evolutionary history in both the geographic distribution and genetic
-structure of their populations. Thanks to the advance of next-generation
+Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In
+this paper, we establish a theoretical framework based on projecting environmental
+Grinellian niches back through time to relate the present geographic distribution of population genetic
+structure to a given species' historical evolutionary context. Thanks to the advance of next-generation
 sequencing technologies, as well as more accurate climate models and an
 enormous amount of information stored in biological collections, it is
-possible to implement this theoretical framework with the help of the
-joint analysis of spatial, environmental and genetic data, with tools
-that are widely distributed among biogeographers and evolutionary
-biologists. Our results reveal that when we have case of isolation by
-distance, suitability prevalence along time corresponds to the fixation
+possible to implement this theoretical framework directly.
+We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species.
+Our results reveal that in cases of genetic isolation by
+geographic distance, suitability prevalence along time corresponds to the genetic fixation
 index of populations with respect to a source population. Populations
 closer to the historical endemic area present a higher genetic diversity
-and a lower \(F_{st}\) value. This framework allows us to add a
+and a lower \(F_{st}\) value. This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a
 biogeographic explanation to the results obtained with population
-genetic methods and to generate maps of this structure as tools to
-support conservation with a perspective that integrates both population
-genetics and historical patterns of species distribution.
+genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to
+support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population
+genetics with historical patterns of species distribution.
 
@@
 
-The current distributions of species do not always correspond to their
-historical distributions over evolutionarily significant periods. This
-is because environmental conditions are not static over time; species
-tend to distribute where conditions are most favorable.
+Current species distributions do not always correspond to their
+historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
+is because environmental conditions are highly dynamic over time, and species
+tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).
 
@@
 thus to indicators of population structure such as the fixation index
-\(F_{st}\). Thus, it is possible to relate patterns of changes in the
+\(F_{st}\). It should therefore be possible to relate patterns of change in the
 distribution of species to the genetic structure of their populations
@@
 
-With this approach we can predict the geographic pattern of population
-structure from environmental information. This approach is strongly
+With this approach we expect to predict the geographic pattern of population-genetic
+structure from abiotic environmental information alone. This approach is strongly
 driven by advances in currently available climate simulations (Leonardi
-et al. 2023; Krapp et al. 2021), as well as next generation sequencing
+et al. 2023; Krapp et al. 2021), as well as next-generation sequencing
 data, and supported by both ecological niche (Thorup et al. 2021;
@@
 Manthey 2014).
+Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.
 
-In this work we propose a method to find the Suitability Prevalence Area
-(SPA) as an index with a double purpose: 1) to find endemic areas to
-delimit the historical distribution of the species and 2) to explain the
-patterns of genetic diversity.
+In this work, we propose a method to find what we call the Suitability Prevalence Area
+(SPA) that is the time-averaged area of highest environmental suitability for a given species.
+The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of the target species; and 2) predicting the expected patterns of genetic diversity given the inferred history of population stability and connectivity.
 
-To obtain the SPA, we performed a historical reconstruction of the
-geographical range back to 120 000 BC at 2 000 year intervals and
-recorded the environmental suitability at each site.
+To obtain the SPA, we overlapped historical reconstructions of the
+geographical distribution of a target species estimated every 2,000 years from the present day until 120,000 years ago, in each time interval recording the environmental suitability at each modeled site.
 
@@
 
-\hypertarget{fixation-index-as-a-functino-of-spa}{%
-\subsubsection{Fixation index as a functino of
+\hypertarget{fixation-index-as-a-function-of-spa}{%
+\subsubsection{Fixation index as a function of
 SPA}\label{fixation-index-as-a-functino-of-spa}}
@@
 (2018). It consideres 10 populations along the known distribution and
-provides values of \(F_st\) between this populations in matrix form. The
+provides values of \(F_{st}\) between this populations in matrix form. The
 matrix was transformed into a table as pairs between populations with
 index values between each pair. The values of the population with
-itself, i.e.~\(0\), were included. Subsequently, georeferences were
+itself, i.e.~\(0\), were included. Subsequently, geografic references were
 obtained for each population considered in the fixation index matrix.
-From this geo-referencing the SPA value for each population was
-extracted. Finally, the SPA value was aligned in the table of fixation
+The SPA value for each population was
+extracted for each population site. Finally, the SPA value was aligned in the table of fixation
 index pairs. In total the table has 4 columns: source population, target
@@
 \endlastfoot
-\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
-\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
-\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
-\emph{S. a. aberti San Juan} & 0 & 0.952 \\
-\emph{S. a. barberi} & 0.794 & 0.001 \\
-\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
-\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
-\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
-\emph{S. a. ferreus Pike} & 0 & 0.967 \\
-\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
+\textit{S. a. aberti Carson-SFW} & 0 & 0.962 \\
+\textit{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
+\textit{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
+\textit{S. a. aberti San Juan} & 0 & 0.952 \\
+\textit{S. a. barberi} & 0.794 & 0.001 \\
+\textit{S. a. chuscensis E} & 0.002 & 0.892 \\
+\textit{S. a. chuscensis W} & 0.018 & 0.713 \\
+\textit{S. a. ferreus Carson E} & 0.04 & 0.581 \\
+\textit{S. a. ferreus Pike} & 0 & 0.967 \\
+\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
 \end{longtable}

### 2023-08-01 manuscript_english.tex
Additional edits to Intro and start of framework

@@
 
-\title{Suitability prevalence area index in late quaternary explains
-genetic diversity in Tassel eared Squirrels}
+\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
 \author{Norma Hernandez, Angel Robles Nathan Upham}
@@
 
-Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In
-this paper, we establish a theoretical framework based on projecting environmental
-Grinellian niches back through time to relate the present geographic distribution of population genetic
-structure to a given species' historical evolutionary context. Thanks to the advance of next-generation
-sequencing technologies, as well as more accurate climate models and an
-enormous amount of information stored in biological collections, it is
-possible to implement this theoretical framework directly.
+Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established.
+In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.
+Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly.
 We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species.
@@
 geographic distance, suitability prevalence along time corresponds to the genetic fixation
-index of populations with respect to a source population. Populations
-closer to the historical endemic area present a higher genetic diversity
-and a lower \(F_{st}\) value. This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a
-biogeographic explanation to the results obtained with population
-genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to
-support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population
-genetics with historical patterns of species distribution.
+index of populations with respect to a source population.
+Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value.
+This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts.
+Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.
 
@@
 In this work, we propose a method to find what we call the Suitability Prevalence Area
-(SPA) that is the time-averaged area of highest environmental suitability for a given species.
-The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of the target species; and 2) predicting the expected patterns of genetic diversity given the inferred history of population stability and connectivity.
+(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period.
+The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.
 
-To obtain the SPA, we overlapped historical reconstructions of the
-geographical distribution of a target species estimated every 2,000 years from the present day until 120,000 years ago, in each time interval recording the environmental suitability at each modeled site.
+As a case study, we consider Tassel-eared Squirrels (\emph{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution (Bono et al.
+2018; Burgin et al.
+2018).
+To obtain the SPA for \textit{S.
+aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.
 
 Subsequently, we delimited historical endemic areas to locations where
-the prevalence of suitability remained at 90\% during this period. From
-the fixation index calculated with respect to populations in the
-historical endemic areas, a statistical model of the fixation index as a
-function of SPA was performed. With this statistical model, the fixation
-index values were projected to the current distribution to obtain a map
-with the geographic pattern of this index.
+the prevalence of suitability remained at 90\% during this period.
+
+Subsequently, we calculated the fixation index (\(F_{st}\)) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA.
+With this statistical model, the \(F_{st}\) values were projected to the current geographic distribution of \textit{S.
+aberti} to obtain a predictive map of expected \(F_{st}\) values over space.
 
@@
 environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
-relates evolutionary fitness to environment, that is, for a species with
+relates evolutionary fitness to environment. That is, for a species with
 non-overlapping generations, the population's net reproductive rate
@@
 
-That is, the sites in the geography \(\mathbf{G}_o\) that present
-positive growth rates are related to favorable conditions. If we take
+In other words, the sites in the geography \(\mathbf{G}_o\) that present
+positive population growth rates are related to favorable environmental conditions. If we take
 the set of places in ambient space where these values are greater than a
-threshold, we can define a volume in multidimensional space such that
-\(R(\vec{e})\) is always positive and greater than this threshold, that
-is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
+certain threshold, we can define a volume in multidimensional space such that
+\(R(\vec{e})\) is always positive and greater than this threshold, as \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
 
-This is a way of expressing the fundamental niche as a function of
-environmental conditions. So we can map each point of geography to
-ambient space. Thus, for each site in the geography there will
+This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a function of
+environmental conditions. Thus, for each site in the geography there will
 correspond a site in the environmental space with different
-environmental conditions and different growth rates. However all sites
+environmental conditions and different growth rates. However, all sites
 within the region defined by \(\mathbf{N}_F\) will have viable
-populations and positive rates. However, we assume that the environment
+populations and positive growth rates. Nonetheless, we assume that the environment
 is dynamic and each geographic point is static, so for each site in
-geographic space this growth rate depends on the environment and
-therefore is different at each time.
+geographic space this growth rate depends on the environmental conditions and
+therefore growth rate changes in value at each time interval.
 
-Thus locally the growth rate will be related to the local population
+Locally, the growth rate will be related to the local population
 size as long as it is within the boundary defined by \(\mathbf{N}_F\).
-Therefore, the local population size as a function of the intrinsic rate
-that depends on the change in environmental conditions over time is:
+Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
 \(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
@@
 
-As a case study we consider squirrels (\emph{Sciurus aberti}), a species
+As a case study, we consider squirrels (\emph{Sciurus aberti}), a species
 currently distributed in disjunct patches from the southern Rocky

### 2023-08-03 manuscript_english.tex
update

@@
 \hypersetup{
-  pdftitle={Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels},
+  pdftitle={Environmental suitability throughout the late Quaternary explains population genetic diversity.},
   pdfauthor={Norma Hernandez, Angel Robles Nathan Upham},
@@
 
-\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
+\title{Environmental suitability throughout the late Quaternary explains
+population genetic diversity.}
 \author{Norma Hernandez, Angel Robles Nathan Upham}
@@
 
-Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established.
-In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.
-Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly.
-We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species.
-Our results reveal that in cases of genetic isolation by
-geographic distance, suitability prevalence along time corresponds to the genetic fixation
-index of populations with respect to a source population.
-Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value.
-This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts.
-Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.
+Genetic variation among populations is known to exhibit biogeographic
+patterns in many species, but general rules of spatial genetic variation
+have not been established. In this paper, we establish a theoretical
+framework based on projecting environmental Grinellian niches back
+through time to relate the present geographic distribution of population
+genetic structure to a given species' historical evolutionary context.
+
+Thanks to the advance of next-generation sequencing technologies, as
+well as more accurate climate models and an enormous amount of
+information stored in biological collections, it is possible to
+implement this theoretical framework directly. We use as a case study
+the Tassel-eared Squirrel (\emph{Sciurus aberti}) to jointly analyze
+spatial, environmental, and genetic data to predict the historical
+endemic area of this species. Our results reveal that in cases of
+genetic isolation by geographic distance, the prevalence of
+environmental suitability over time corresponds to the genetic fixation
+index of populations with respect to a source population. Populations
+closer to the historical endemic area present a higher genetic diversity
+and a lower\(F_{st}\) value.
+
+This empirical example fits into a theoretical framework, which allows
+us two advances: (i) we can now add a biogeographic explanation to the
+results obtained with population genetic methods; and (ii) we can
+generate maps of this genetic structure as predictive tools to support
+biodiversity conservation efforts. Overall, this work advances a
+perspective that integrates population genetics with historical patterns
+of species distribution.
 
@@
 Current species distributions do not always correspond to their
-historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
-is because environmental conditions are highly dynamic over time, and species
-tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).
+historical distributions over evolutionarily significant periods. Often
+they differ substantially, such as in the case of postglacial range
+expansion in high latitude environments (e.g., Lessa, Cook, and Patton
+(2003)). This is because environmental conditions are highly dynamic
+over time, and species tend to track those environments, distributing
+where conditions are most favorable. That is, species environmental
+niches tend to be conserved (Peterson (2011)).
 
@@
 thus to indicators of population structure such as the fixation index
-\(F_{st}\). It should therefore be possible to relate patterns of change in the
-distribution of species to the genetic structure of their populations
-using a statistical model that explains this relationship.
+\(F_{st}\). It should therefore be possible to relate patterns of change
+in the distribution of species to the genetic structure of their
+populations using a statistical model that explains this relationship.
 
-With this approach we expect to predict the geographic pattern of population-genetic
-structure from abiotic environmental information alone. This approach is strongly
-driven by advances in currently available climate simulations (Leonardi
-et al. 2023; Krapp et al. 2021), as well as next-generation sequencing
-data, and supported by both ecological niche (Thorup et al. 2021;
-Nogués-Bravo 2009) and population genetics theories (Lira-Noriega and
-Manthey 2014).
-Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.
+With this approach we expect to predict the geographic pattern of
+population-genetic structure from abiotic environmental information
+alone. This approach is strongly driven by advances in currently
+available climate simulations (Leonardi et al. 2023; Krapp et al. 2021),
+as well as next-generation sequencing data, and supported by both
+ecological niche (Thorup et al. 2021; Nogués-Bravo 2009; Lira-Noriega
+and Manthey 2014) and population genetics theories (Walsh and Lynch
+2018). Importantly, these predicted patterns of population structure
+will serve as a type of \textit{a priori} hypothesis for the influence
+of abiotic factors alone, with the empirical detection of deviations
+from this hypothesis indicating that biotic factors such as competition
+or predation from other species have also impacted the observed
+population structure.
 
-In this work, we propose a method to find what we call the Suitability Prevalence Area
-(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period.
-The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.
+In this work, we propose a method to find an index that we call the
+Suitability Prevalence Area (SPA) that is the area of highest
+environmental suitability for a given species over a historical time
+period. The SPA is an index with a double purpose: 1) identifying areas
+of greatest likelihood for the historical distribution of a target
+species (i.e., historical endemism); and 2) to explain the expected
+patterns of genetic diversity given the species' inferred history of
+population stability and connectivity.
 
-As a case study, we consider Tassel-eared Squirrels (\emph{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution (Bono et al.
-2018; Burgin et al.
-2018).
-To obtain the SPA for \textit{S.
-aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.
+As a case study, we consider Tassel-eared Squirrels
+(\textit{Sciurus aberti}), a species of rodent that is currently
+distributed in disjunct patches from the southern Rocky Mountains in the
+United States to the northern Sierra Madre Occidental in Mexico for
+which it is possible to find reliable information on both the genetic
+structure of its populations and its current distribution (Bono et al.
+2018; Burgin et al. 2018). To obtain the SPA for \textit{S. aberti}, we
+overlapped historical reconstructions of the species' geographical
+distribution estimated every 2,000 years from the present day until
+120,000 years ago, recording the environmental suitability in each time
+interval at each modeled site.
 
-Subsequently, we delimited historical endemic areas to locations where
-the prevalence of suitability remained at 90\% during this period.
-
-Subsequently, we calculated the fixation index (\(F_{st}\)) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA.
-With this statistical model, the \(F_{st}\) values were projected to the current geographic distribution of \textit{S.
-aberti} to obtain a predictive map of expected \(F_{st}\) values over space.
+Subsequently, we calculated the fixation index (\(F_{st}\)) of the
+genetic relatedness among populations and compared that to the SPA index
+of historical endemism, constructing a statistical model of the fixation
+index as a function of the SPA. With this explanatory model, the
+\(F_{st}\) values were projected to the current geographic distribution
+of \textit{S. aberti} to obtain a predictive map of expected \(F_{st}\)
+values over space.
 
@@
 
-In other words, the sites in the geography \(\mathbf{G}_o\) that present
-positive population growth rates are related to favorable environmental conditions. If we take
-the set of places in ambient space where these values are greater than a
-certain threshold, we can define a volume in multidimensional space such that
-\(R(\vec{e})\) is always positive and greater than this threshold, as \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
+In other words, \(\mathbf{G}_o\) is the set of sites in the geography
+that present positive growth rates related to favorable conditions. If
+we take the set of places in environmental space where these values are
+greater than a threshold, we can define a volume in multidimensional
+space such that \(R(\vec{e})\) is always positive and greater than this
+threshold, that is
+\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
 
-This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a function of
-environmental conditions. Thus, for each site in the geography there will
-correspond a site in the environmental space with different
-environmental conditions and different growth rates. However, all sites
-within the region defined by \(\mathbf{N}_F\) will have viable
-populations and positive growth rates. Nonetheless, we assume that the environment
-is dynamic and each geographic point is static, so for each site in
-geographic space this growth rate depends on the environmental conditions and
-therefore growth rate changes in value at each time interval.
+This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a
+function of environmental conditions. Thus, for each site in the
+geography there will correspond a site in the environmental space with
+different environmental conditions and different growth rates. However,
+all sites within the region defined by \(\mathbf{N}_F\) will have viable
+populations and positive growth rates. Nonetheless, we assume that the
+environment is dynamic and each geographic point is static, so for each
+site in geographic space this growth rate depends on the environmental
+conditions and therefore growth rate changes in value at each time
+interval.
 
-Locally, the growth rate will be related to the local population
-size as long as it is within the boundary defined by \(\mathbf{N}_F\).
-Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
-\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
-population size is a function of ideal environmental conditions.
+Locally, the growth rate will be related to the local population size as
+long as it is within the boundary defined by \(\mathbf{N}_F\).
+Therefore, the local population size as a function of the intrinsic
+growth rate, which depends on the change in environmental conditions
+over time, can be expressed as: \(N_i(t+1) = R(\vec{e_i}(t) ) N_i(t)\).
+That is, the locally population size is a function of ideal
+environmental conditions.
 
@@
 the population grows, that is,
-\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
-any local population will have a positive growth rate.
+\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\) and therefore any
+local population will have a positive growth rate.
 
@@
 
-This equation was proposed by Soberon and Peterson Soberón and Peterson
-(2020) to relate demographics to a function of evolutionary fitness and
-environment. In that study they tested the validity with monthly
-variables and empirical values on a short time scale. relating life
-history parameters and the environmental suitability of niche models.
+This equation was proposed by Soberon and Peterson (2020) to relate
+demographics to a function of evolutionary fitness and environment. In
+that study they tested the validity with monthly variables and empirical
+values on a short time scale. relating life history parameters and the
+environmental suitability of niche models.
 
@@
 relate this growth rate to an environmental suitability function bounded
-between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
-it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any
-function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
-\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
-of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
-approximated from points of presence in the geography obtained from
-databases, numerically approximating a convex surface with a surface
-\(\mathbf{N}_R\) of the environmental space.
+between \(0\) and \(1\), for example, for a unimodal multivariate
+probability density function, it would be equivalent to the distance
+from the maximum of the function. That is, the fitness-related
+environmental suitability can be any function such that it is \(0\) if
+\(R_i(\vec{e}_i) > 1\) and \(1\) when \(R\) is evaluated at the expected
+value of \(\vec {e}\) for all values of \(R_i(\vec{e}_i) > 1\). In a
+practical way, \(\mathbf{N}_F\) is approximated from points of presence
+in the geography obtained from databases, numerically approximating a
+convex surface with a surface \(\mathbf{N}_R\) of the environmental
+space.
 
@@
 The goal of this theoretical approach is not to know the shape of the
-function of \(R_0\) but rather to find this local variation relating
-this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
-\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
-estimated from known methods of applied mathematics and machine learning
-(for example the case of Maxent, Phillips, Anderson, and Schapire
-(2006)), i.e.~:
+function of \(R\) but rather to find this local variation relating this
+value of \(R\) and the function \(\mathbf{N}_F\) with a function \$
+S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be estimated
+from known methods of applied mathematics and machine learning (for
+example the case of Maxent, (Phillips, Anderson, and Schapire 2006)),
+i.e.~:
 
@@
 where if suitability is \(0\) the population is \(0\) and if suitability
-is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)
+is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) in
+some point of the environmental space.
 
@@
 effective population size \(N_e\) when subpopulations diverge from a
-common ancestor in \(t\) generations. Wright (1943):
+common ancestor in \(t\) generations. Walsh and Lynch (2018):
 
@@
 that although the number of generations is not exactly equal to the time
-of the fixation index, they are expected to be very close.
+of the fixation index, they are expected to be very close:
 
@@
 where the \(F_{st}\) is lower are in areas where the prevalence over
-time is higher, i.e.~environmental conditions remain suitable for a
-significant period of time and therefore no effects of genetic drift of
-populations have occurred. Conversely, it is expected that in
-populations with a high fixation index value, suitability will not have
-prevailed over the same time interval even though conditions currently
-exist and populations are present.
+time is higher, i.e.\textasciitilde environmental conditions remain
+suitable for a significant period of time and therefore no effects of
+genetic drift of populations have occurred. Conversely, it is expected
+that in populations with a high fixation index value, suitability will
+not have prevailed over the same time interval even though conditions
+currently exist and populations are present.
 
@@
 \item
-  A linear model of the fixation index value with respect to the SPA is
-  performed.
+  Perform a linear model of the fixation index value with respect to the
+  SPA.
 \item
@@
 \item
-  A map is generated with all the values interpolated by the model.
+  Generate a map with all the values interpolated by the model.
 \end{itemize}
 
-As a case study, we consider squirrels (\emph{Sciurus aberti}), a species
+As a case study we consider squirrels (\emph{Sciurus aberti}), a species
 currently distributed in disjunct patches from the southern Rocky
@@
 we list the following considerations:
+
+As a case study, we consider squirrels (\emph{Sciurus aberti}), a
+species currently distributed in disjunct patches from the southern
+Rocky Mountains in the United States to the northern Sierra Madre
+Occidental in Mexico for which it is possible to find reliable
+information on both the genetic structure of its populations and its
+current distribution (Bono et al. 2018; Burgin et al. 2018). In order to
+select this species, we list the following considerations:
 
@@
 
-\hypertarget{analogy-of-environmental-conditions}{%
-\paragraph{Analogy of environmental
-conditions}\label{analogy-of-environmental-conditions}}
-
-To be certain that areas in the past are climatically analogous to the
-present we implemented the mobility-orient parity (MOP) test. This
-method generates a value from \(0\) to \(1\) according to the
-environmental analogy of past conditions to the present to eliminate
-geographic areas in the projections to the past outside of climates
-analogous to the present (Owens et al. 2013).
-
 \hypertarget{suitability-prevalence-area-spa}{%
@@
 \subsubsection{Fixation index as a function of
-SPA}\label{fixation-index-as-a-functino-of-spa}}
+SPA}\label{fixation-index-as-a-function-of-spa}}
 
@@
 (2018). It consideres 10 populations along the known distribution and
-provides values of \(F_{st}\) between this populations in matrix form. The
+provides values of \(F_st\) between this populations in matrix form. The
 matrix was transformed into a table as pairs between populations with
 index values between each pair. The values of the population with
-itself, i.e.~\(0\), were included. Subsequently, geografic references were
+itself, i.e.~\(0\), were included. Subsequently, georeferences were
 obtained for each population considered in the fixation index matrix.
-The SPA value for each population was
-extracted for each population site. Finally, the SPA value was aligned in the table of fixation
+From this geo-referencing the SPA value for each population was
+extracted. Finally, the SPA value was aligned in the table of fixation
 index pairs. In total the table has 4 columns: source population, target
@@
 \endlastfoot
-\textit{S. a. aberti Carson-SFW} & 0 & 0.962 \\
-\textit{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
-\textit{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
-\textit{S. a. aberti San Juan} & 0 & 0.952 \\
-\textit{S. a. barberi} & 0.794 & 0.001 \\
-\textit{S. a. chuscensis E} & 0.002 & 0.892 \\
-\textit{S. a. chuscensis W} & 0.018 & 0.713 \\
-\textit{S. a. ferreus Carson E} & 0.04 & 0.581 \\
-\textit{S. a. ferreus Pike} & 0 & 0.967 \\
-\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
+\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
+\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
+\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
+\emph{S. a. aberti San Juan} & 0 & 0.952 \\
+\emph{S. a. barberi} & 0.794 & 0.001 \\
+\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
+\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
+\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
+\emph{S. a. ferreus Pike} & 0 & 0.967 \\
+\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
 \end{longtable}
@@
 species can potentially be found but share different degrees of genetic
-diversity.
+diversity..
 
 Projections of genetic diversity in geography have been made in previous
-studies Van Zonneveld et al. (2012). However in a general way they are
+studies (Van Zonneveld et al. 2012). However in a general way they are
 extrapolations of observations, without a predictive model, which with
@@
 paper, it is possible to explain the explanatory value of the
-Coefficient of determination for our linear model (\$R\^{}2 = 0.79 \$),
-which is information coming from two completely different sources,
-without any correlation, but always guided by the premise of the
-evolutionary process as causal.
+coefficient of determination for our linear model, \(R^{2} = 0.79\),
+which is information coming from two completely different sources, but
+always guided by the premise of the evolutionary process as causal.
 
@@
 studies with available information, as well as the design of new
-questions, e.g.~species assemblages under prevalence of environmental
-suitability. A research path was also established on whether the process
-of environmental stability over time is related to biodiversity.
+questions, e.g.\textasciitilde species assemblages under prevalence of
+environmental suitability. A research path was also established on
+whether the process of environmental stability over time is related to
+biodiversity.
+
+It is also noteworthy to mention that in previous taxonomic
+descriptions, it was taken into account that the distribution of Sciurus
+aberti was only related to the distribution of Arizona pinus ponderosa.
+However, there is a consensus on the range of distribution of this pine,
+which is limited to the southern United States (Hess and Fulé 2020).
+Therefore the known distribution of Sciurus aberti is in mixed
+coniferous forests. This plasticity in the squirrel-pine interaction
+explains the expansion of the species in interglacial periods and its
+distribution in warmer periods, such as the current one. However, the
+low genetic diversity found in places where the squirrel species
+interact with few pine species could put it at greater risk of climate
+change that constricts the distribution of p.~p.~Arizona, especially in
+the Sky Island region of Arizona.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hess2020mexican}{}}%
+Hess, Victoria A, and Peter Z Fulé. 2020. {``Is a Mexican Pine Species
+Better Adapted to the Warming Climate of the Southwestern USA?''}
+\emph{Frontiers in Forests and Global Change} 3: 60.
+
 \leavevmode\vadjust pre{\hypertarget{ref-karlin1968rates}{}}%
@@
 Use Paleoclimatic Reconstructions.''} \emph{Ecography}, e06481.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lessa2003genetic}{}}%
+Lessa, Enrique P, Joseph A Cook, and James L Patton. 2003. {``Genetic
+Footprints of Demographic Expansion in North America, but Not Amazonia,
+During the Late Quaternary.''} \emph{Proceedings of the National Academy
+of Sciences} 100 (18): 10331--34.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-owens2013constraints}{}}%
-Owens, Hannah L, Lindsay P Campbell, L Lynnette Dornak, Erin E Saupe,
-Narayani Barve, Jorge Soberón, Kate Ingenloff, et al. 2013.
-{``Constraints on Interpretation of Ecological Niche Models by Limited
-Environmental Ranges on Calibration Areas.''} \emph{Ecological
-Modelling} 263: 10--18.
+\leavevmode\vadjust pre{\hypertarget{ref-peterson2011ecological}{}}%
+Peterson, A Townsend. 2011. {``Ecological Niche Conservatism: A
+Time-Structured Review of Evidence.''} \emph{Journal of Biogeography} 38
+(5): 817--27.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lynchi2018}{}}%
+Walsh, Bruce, and Michael Lynch. 2018. {``{59The Genetic Effective Size
+of a Population}.''} In \emph{{Evolution and Selection of Quantitative
+Traits}}. Oxford University Press.
+\url{https://doi.org/10.1093/oso/9780198830870.003.0003}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%

### 2023-08-03 manuscript_english.tex
Updates from Overleaf

--- 2023-08-01 manuscript_english.tex
@@
 
-\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
+\title{Environmental suitability throughout the late Quaternary explains population genetic diversity}
 \author{Norma Hernandez, Angel Robles Nathan Upham}
@@
 In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.
+
 Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly.
-We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species.
-Our results reveal that in cases of genetic isolation by
-geographic distance, suitability prevalence along time corresponds to the genetic fixation
-index of populations with respect to a source population.
+We use as a case study the Tassel-eared Squirrel (\textit{Sciurus aberti}) to jointly analyze spatial, environmental, and genetic data to predict the historical endemic area of this species.
+Our results reveal that in cases of genetic isolation by geographic distance, the prevalence of environmental suitability over time corresponds to the genetic fixation index of populations with respect to a source population.
 Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value.
+
 This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts.
@@
 Current species distributions do not always correspond to their
-historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
-is because environmental conditions are highly dynamic over time, and species
-tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).
+historical distributions over evolutionarily significant periods.
+Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS).
+This is because environmental conditions are highly dynamic over time, and speciestend  to track those environments, distributing where conditions are most favorable.
+That is, species environmental niches tend to be conserved (REF).
 
@@
 
-In this work, we propose a method to find what we call the Suitability Prevalence Area
-(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period.
-The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.
+In this work, we propose a method to find an index that we call the Suitability Prevalence Area
+(SPA) that is the area of highest environmental suitability for a given species over a historical time period.
+The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) to explain the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.
 
@@
 aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.
-
-Subsequently, we delimited historical endemic areas to locations where
-the prevalence of suitability remained at 90\% during this period.
 
@@
 Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
-\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
-population size is a function of ideal environmental conditions.
+\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally population size is a function of ideal environmental conditions.
 
@@
 
-This equation was proposed by Soberon and Peterson Soberón and Peterson
-(2020) to relate demographics to a function of evolutionary fitness and
+This equation was proposed by Soberon and Peterson (2020) to relate demographics to a function of evolutionary fitness and
 environment. In that study they tested the validity with monthly
@@
 The goal of this theoretical approach is not to know the shape of the
-function of \(R_0\) but rather to find this local variation relating
-this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
+function of \(R\) but rather to find this local variation relating
+this value of \(R\) and the function \(\mathbf{N}_F\) with a function
 \$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
@@
 i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
-generation time), as we describe above, so we can asume an harmonic mean
+generation time), as we describe above, so we can assume an harmonic mean
 of suitabilities
@@
 F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
-\] If we consider all sites in the available geography, we call this
+\]
+
+If we consider all sites in the available geography, we call this
 harmonic mean \emph{Suitability Prevalence Area}
@@
 
+Other topics to cover
+- Dependency on \textit{Pinus ponderosa} -- is in doubt, P.
+p.
+arizonica in the USA, but in the Sierra Madre is a different pine species -- so there is likely more lability in the environmental niche than expected -- less of a Ponderosa specialist than assumed -- also the fungi that is associated with pines
+
 \hypertarget{references}{%

### 2023-08-08 manuscript_english.Rmd

@@
 ---
-title: Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels
+title: Environmental suitability throughout the late Quaternary explains population genetic diversity. 
 author: "Norma Hernandez, Angel Robles Nathan Upham"
@@
 
-Genetic variation among populations exhibits biogeographic patterns.
-In this paper we establish a theoretical framework based on environmental Grinellian niches as a tool for modeling species distributions, as well as their projected past environmental scenarios to relate the present evolutionary history in both the geographic distribution and genetic structure of their populations.
-Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework with the help of the joint analysis of spatial, environmental and genetic data, with tools that are widely distributed among biogeographers and evolutionary biologists.
-Our results reveal that when we have case of isolation by distance, suitability prevalence along time corresponds to the fixation index of populations with respect to a source population.
-Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value.
-This framework allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.
+Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established.
+In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.
+
+Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly.
+We use as a case study the Tassel-eared Squirrel (*Sciurus aberti*) to jointly analyze spatial, environmental, and genetic data to predict the historical endemic area of this species.
+Our results reveal that in cases of genetic isolation by geographic distance, the prevalence of environmental suitability over time corresponds to the genetic fixation index of populations with respect to a source population.
+Populations closer to the historical endemic area present a higher genetic diversity and a lower$F_{st}$ value.
+
+This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts.
+Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.
 
@@
 
-The current distributions of species do not always correspond to their historical distributions over evolutionarily significant periods.
-This is because environmental conditions are not static over time; species tend to distribute where conditions are most favorable.
+Current species distributions do not always correspond to their
+historical distributions over evolutionarily significant periods.
+Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., @lessa2003genetic).
+This is because environmental conditions are highly dynamic over time, and species
+tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (@peterson2011ecological).
 
 It follows from this relationship between environment and species distribution that population size also varies with time.
-This variation in population size is related to the effective population size $N_e$ [@karlin1968rates], so that an index reflecting changes in environmental conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index $F_{st}$.
-Thus, it is possible to relate patterns of changes in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.
+This variation in population size is related to the effective population size $N_e$
+[@karlin1968rates], so that an index reflecting changes in environmental
+conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index
+$F_{st}$.
+It should therefore be possible to relate patterns of change in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.
 
-With this approach we can predict the geographic pattern of population structure from environmental information.
-This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting] and population genetics theories [@lira2014relationship].
+With this approach we expect to predict the geographic pattern of population-genetic structure from abiotic environmental information alone.
+This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next-generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting; @lira2014relationship] and population genetics theories [@lynchi2018].
+Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.
 
-In this work we propose a method to find the Suitability Prevalence Area (SPA) as an index with a double purpose: 1) to find endemic areas to delimit the historical distribution of the species and 2) to explain the patterns of genetic diversity.
+In this work, we propose a method to find an index that we call the Suitability Prevalence Area (SPA) that is the area of highest environmental suitability for a given species over a historical time period.
+The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) to explain the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.
 
-To obtain the SPA, we performed a historical reconstruction of the geographical range back to 120 000 BC at 2 000 year intervals and recorded the environmental suitability at each site.
+As a case study, we consider Tassel-eared Squirrels (\textit{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].
+To obtain the SPA for \textit{S.
+aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.
 
-Subsequently, we delimited historical endemic areas to locations where the prevalence of suitability remained at 90% during this period.
-From the fixation index calculated with respect to populations in the historical endemic areas, a statistical model of the fixation index as a function of SPA was performed.
-With this statistical model, the fixation index values were projected to the current distribution to obtain a map with the geographic pattern of this index.
+Subsequently, we calculated the fixation index ($F_{st}$) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA.
+With this explanatory model, the $F_{st}$ values were projected to the current geographic distribution of \textit{S.
+aberti} to obtain a predictive map of expected $F_{st}$ values over space.
 
@@
 
-We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables.
-Then $R_i(\vec{e}_i)$ is a function that relates evolutionary fitness to environment,
-that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of evolutionary fitness that depends on environmental conditions.
+We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
+environmental variables, that is, a point in the space of \(v\)
+environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
+relates evolutionary fitness to environment. That is, for a species with
+non-overlapping generations, the population's net reproductive rate 
+\(R_i\) is a measure of evolutionary fitness that depends on
+environmental conditions.
 
-That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
-If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
+In other words, $\mathbf{G}_o$ is the set of sites in the geography that present positive growth rates related to favorable conditions.
+If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
 
-This is a way of expressing the fundamental niche as a function of environmental conditions.
-So we can map each point of geography to ambient space.
-Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates.
-However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates.
-However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.
+This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a function of environmental conditions. Thus, for each site in the geography there will
+correspond a site in the environmental space with different
+environmental conditions and different growth rates. However, all sites
+within the region defined by \(\mathbf{N}_F\) will have viable
+populations and positive growth rates. Nonetheless, we assume that the environment is dynamic and each geographic point is static, so for each site in
+geographic space this growth rate depends on the environmental conditions and
+therefore growth rate changes in value at each time interval.
 
-Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.
-Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
-$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.
+Locally, the growth rate will be related to the local population
+size as long as it is within the boundary defined by \(\mathbf{N}_F\).
+Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
+\(N_i(t+1) = R(\vec{e_i}(t) ) N_i(t)\). That is, the locally population size is a function of ideal environmental conditions.
 
-We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space.
-Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.
+We start by considering this defined volume of the set of \(R_0\) as a
+set of trajectories for each point from geographic space to
+environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
+the population grows, that is,
+\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\) and therefore
+any local population will have a positive growth rate.
 
@@
 
-This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of evolutionary fitness and environment.
+This equation was proposed by Soberon and Peterson [-@soberon2020shape] to relate demographics to a function of evolutionary fitness and environment.
 In that study they tested the validity with monthly variables and empirical values on a short time scale.
@@
 
-This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function.
-That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$.
-In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.
+This set can be seen as a convex volume in environmental space and we
+relate this growth rate to an environmental suitability function bounded
+between \(0\) and \(1\), for example, for a unimodal multivariate probability density function,
+it would be equivalent to the distance from the maximum of the function.
+That is, the fitness-related environmental suitability can be any
+function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
+\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
+of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
+approximated from points of presence in the geography obtained from
+databases, numerically approximating a convex surface with a surface
+\(\mathbf{N}_R\) of the environmental space.
 
-Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory, the fitness function changes and, therefore, the suitability function changes but not accordingly, in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.
+Once we have an approximation of this volume, we consider that each
+point in the geography is static and has a trajectory in the
+environmental space and that for each trajectory, the fitness function
+changes and, therefore, the suitability function changes but not
+accordingly, in a homogeneous way, if not more so that each site
+presents different displacement rates and different demographic rates.
 
-The goal of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum), i.e.
+The goal of this theoretical approach is not to know the shape of the function of $R$ but rather to find this local variation relating this value of $R$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, [@phillips2006maximum]), i.e.
 :
@@
 $$
-Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
+Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$ in some point of the environmental space.
 
@@
 We start from the assumption that in the absence of gene flow between subpopulations, the rate of fixation by isolation depends on the effective population size $N_e$ when subpopulations diverge from a common ancestor in $t$ generations.
-[@hedrick2005standardized: @wright1943isolation]:
+[@hedrick2005standardized: @wright1943isolation, @lynchi2018]:
 
@@
 + \frac{1}{N_\tau} } \Big)$.
-We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close.
+We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close:
 
@@
 $$
-If we consider all sites in the available geography, we call this harmonic mean *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ (SPA) since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e.
-environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred.
-Conversely, it is expected that in populations with a high fixation index value, suitability will not have prevailed over the same time interval even though conditions currently exist and populations are present.
+If we consider all sites in the available geography, we call this
+harmonic mean \emph{Suitability Prevalence Area}
+\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
+where the \(F_{st}\) is lower are in areas where the prevalence over
+time is higher, i.e.~environmental conditions remain suitable for a
+significant period of time and therefore no effects of genetic drift of
+populations have occurred. Conversely, it is expected that in
+populations with a high fixation index value, suitability will not have
+prevailed over the same time interval even though conditions currently
+exist and populations are present.
 
@@
 
-1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
+1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
 2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
@@
 - Extract values from the SPA map with these points. 
-- A linear model of the fixation index value with respect to the SPA is performed.
+- Perform a linear model of the fixation index value with respect to the SPA.
 - With the linear model all SPA values of suitability are interpolated.
-- A map is generated with all the values interpolated by the model.
+- Generate a map with all the values interpolated by the model.
 
@@
 In order to select this species, we list the following considerations:
+
+As a case study, we consider squirrels (\emph{Sciurus aberti}), a species
+currently distributed in disjunct patches from the southern Rocky
+Mountains in the United States to the northern Sierra Madre Occidental
+in Mexico for which it is possible to find reliable information on both
+the genetic structure of its populations and its current distribution
+[@bono2018genome; @burgin2018many]. In order to select this species,
+we list the following considerations:
 
@@
 
-#### Analogy of environmental conditions 
-To be certain that areas in the past are climatically analogous to the present we implemented the mobility-orient parity (MOP) test.
-This method generates a value from $0$ to $1$ according to the environmental analogy of past conditions to the present to eliminate geographic areas in the projections to the past outside of climates analogous to the present [@owens2013constraints].
-
 ### Suitability Prevalence Area (SPA)
@@
 
-### Fixation index as a functino of SPA
+### Fixation index as a function of SPA
 
@@
 
-With the current availability of occurrence data and models of potential distribution projected to past climate scenarios it is possible to find geographic patterns of differentiation in populations at an acceptable resolution (half a degree) that allows us to define sites where the species can potentially be found but share different degrees of genetic diversity.
+With the current availability of occurrence data and models of potential
+distribution projected to past climate scenarios it is possible to find
+geographic patterns of differentiation in populations at an acceptable
+resolution (half a degree) that allows us to define sites where the
+species can potentially be found but share different degrees of genetic
+diversity..
 
-Projections of genetic diversity in geography have been made in previous studies @van2012mapping.
-However in a general way they are extrapolations of observations, without a predictive model, which with some explanatory variable behind can explain in a statistically convincing way the information obtained from population genetics.
+Projections of genetic diversity in geography have been made in previous
+studies [@van2012mapping]. However in a general way they are
+extrapolations of observations, without a predictive model, which with
+some explanatory variable behind can explain in a statistically
+convincing way the information obtained from population genetics.
 
-We have shown a generalizable protocol for predicting patterns of population genetics in geography.
-In this way we link information on the evolutionary history of a taxon and its ecology.
-In addition, we provide a predictive vehicle that generates new geographic information applicable to decision making in both survey conduct and taxon conservation.
+We have shown a generalizable protocol for predicting patterns of
+population genetics in geography. In this way we link information on the
+evolutionary history of a taxon and its ecology. In addition, we provide
+a predictive vehicle that generates new geographic information
+applicable to decision making in both survey conduct and taxon
+conservation.
 
-Thanks to the theoretical support that we have briefly outlined in this paper, it is possible to explain the explanatory value of the Coefficient of determination for our linear model ($R^2 = 0.79 $), which is information coming from two completely different sources, without any correlation, but always guided by the premise of the evolutionary process as causal.
+Thanks to the theoretical support that we have briefly outlined in this
+paper, it is possible to explain the explanatory value of the
+coefficient of determination for our linear model, $R^{2} = 0.79$,
+which is information coming from two completely different sources, but always guided by the premise of the evolutionary process as causal.
 
 These maps represent a turning point in the conception of conservation.
-With the tool we propose in this study, it is possible to delimit conservation sites based on evidence of current and historical distribution and genetic diversity among different populations.
+With the tool we propose in this study, it is possible to delimit
+conservation sites based on evidence of current and historical
+distribution and genetic diversity among different populations.
 
-While many conservation decisions are made on the basis of genetic diversity, studying the genetic structure of populations at a macro-scale is often very complicated, and making conservation decisions based on current distribution without taking into account the historical distribution of taxa can lead to a misguided conservation strategy.
+While many conservation decisions are made on the basis of genetic
+diversity, studying the genetic structure of populations at a
+macro-scale is often very complicated, and making conservation decisions
+based on current distribution without taking into account the historical
+distribution of taxa can lead to a misguided conservation strategy.
 
-The generalization of our protocol and its implementation in tools widely used by biogeography allow its direct implementation in many case studies with available information, as well as the design of new questions, e.g.
-species assemblages under prevalence of environmental suitability.
-A research path was also established on whether the process of environmental stability over time is related to biodiversity.
+The generalization of our protocol and its implementation in tools
+widely used by biogeography allow its direct implementation in many case
+studies with available information, as well as the design of new
+questions, e.g.~species assemblages under prevalence of environmental
+suitability. A research path was also established on whether the process
+of environmental stability over time is related to biodiversity.
+
+It is also noteworthy to mention that in previous taxonomic descriptions, it was taken into account that the distribution of Sciurus aberti was only related to the distribution of Arizona pinus ponderosa.
+However, there is a consensus on the range of distribution of this pine, which is limited to the southern United States [@hess2020mexican].
+Therefore the known distribution of Sciurus aberti is in mixed coniferous forests.
+This plasticity in the squirrel-pine interaction explains the expansion of the species in interglacial periods and its distribution in warmer periods, such as the current one.
+However, the low genetic diversity found in places where the squirrel species interact with few pine species could put it at greater risk of climate change that constricts the distribution of p.
+p.
+Arizona, especially in the Sky Island region of Arizona.
 

### 2023-08-08 manuscript_english.Rmd
fix R_0

@@
 In other words, $\mathbf{G}_o$ is the set of sites in the geography that present positive growth rates related to favorable conditions.
-If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.
+If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > k \}$.
 
@@
 
-We start by considering this defined volume of the set of \(R_0\) as a
+We start by considering this defined volume of the set of \(R\) as a
 set of trajectories for each point from geographic space to
-environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
+environmental space. Considering this rate fixed \(\mathbf{R}\) when
 the population grows, that is,

### 2023-08-08 manuscript_english.tex
fix R_0

--- 2023-08-03 manuscript_english.tex
@@
 threshold, that is
-\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).
+\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > k \}\).
 
@@
 
-We start by considering this defined volume of the set of \(R_0\) as a
-set of trajectories for each point from geographic space to
-environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
-the population grows, that is,
-\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\) and therefore any
-local population will have a positive growth rate.
+We start by considering this defined volume of the set of \(R\) as a set
+of trajectories for each point from geographic space to environmental
+space. Considering this rate fixed \(\mathbf{R}\) when the population
+grows, that is, \(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\)
+and therefore any local population will have a positive growth rate.
 
